# Sequences and Series {#sequences}

Business life runs in discrete steps: monthly repayments, annual pay rises, quarterly sales, yearly depreciation. Sequences describe quantities that evolve step by step, and series add them up. This chapter rebuilds the A-level toolkit — recurrence relations, sigma notation, arithmetic and geometric sequences, the binomial expansion — and connects each piece to the financial and operational models where it earns its living.

**Why this matters:** Compound interest, loan repayment, depreciation, cumulative sales and inventory dynamics are all sequence models. The geometric series in particular is the mathematical engine behind almost every formula in finance — if you take a finance module, you will meet it in the first fortnight.

## Sequences and recurrence relations {#recurrence}

### Key idea

A sequence can be defined in two ways:

- **By a formula for the $n$th term**, e.g. $u_n = 3n + 2$, which lets you jump straight to any term.
- **Recursively**, by a rule of the form $u_{n+1} = f(u_n)$ together with a starting value, e.g. $u_{n+1} = 2u_n - 3$ with $u_1 = 4$. Each term is built from the one before — you must iterate your way forward.

Recursive definitions are how dynamic systems are naturally described: *next month's balance is this month's balance, plus interest, minus the repayment*.

### Worked examples

**Example (simple).** The sequence $u_n = 3n + 2$: write down the first three terms and the 50th term.

$u_1 = 5$, $u_2 = 8$, $u_3 = 11$; jumping ahead, $u_{50} = 152$.

**Example (moderate).** A sequence is defined by $u_{n+1} = 2u_n - 3$ with $u_1 = 4$. Find $u_2$, $u_3$ and $u_4$.

$u_2 = 2(4) - 3 = 5$, $u_3 = 2(5) - 3 = 7$, $u_4 = 2(7) - 3 = 11$. No shortcut here — recursion means stepping through.

**Example (challenging — a savings account as a recurrence).** An account pays 2% interest per year, credited at the end of each year, and the saver deposits £500 at the start of each year. Let $a_n$ be the balance just after the $n$th year's interest. Explain why

$$a_{n+1} = 1.02(a_n + 500), \qquad a_0 = 0,$$

and compute the balance after three years.

Each year: take the current balance, add the new £500 deposit, then grow the whole lot by 2% (multiplying by $1.02$). Iterating:

$$a_1 = 1.02(0 + 500) = 510, \quad a_2 = 1.02(510 + 500) = 1030.20, \quad a_3 = 1.02(1030.20 + 500) = 1560.80.$$

Note the saver has paid in £1,500 and holds £1,560.80 — the £60.80 difference is compound interest at work. Later in the chapter, geometric series will give us a *closed formula* for $a_n$, so we won't have to iterate at all.

### Practice exercises

1. For $u_n = 5n - 2$, write down the first three terms and $u_{12}$.
2. A sequence satisfies $u_{n+1} = 3u_n - 4$. Compute $u_2, u_3, u_4$ (a) starting from $u_1 = 2$, (b) starting from $u_1 = 3$. What do you notice in case (a)?
3. An account satisfies $a_{n+1} = 1.02a_n + 50$ with $a_0 = 500$ (interest then a £50 deposit). Find $a_1$, $a_2$ and $a_3$ to the nearest penny.

**Common mistakes:**
- Confusing the two kinds of definition: $u_n = 3n + 2$ gives any term directly; $u_{n+1} = 3u_n + 2$ does not — check which you've been given before "jumping ahead".
- Losing track of where the sequence starts: $u_0$ and $u_1$ conventions both exist. State which you're using.

**Why this matters:** In exercise 2(a) you found a *fixed point* — a starting value the rule never moves. Fixed points of recurrences are exactly what the iterative methods of Chapter \@ref(numerical) hunt for, and "steady states" of business systems (stable inventory levels, equilibrium market shares) are fixed points too.

::: {.webex-check .webex-box}
**Self-check: sequences and recurrence relations**

**Q1.** For the sequence $u_n = 4n - 1$, find $u_{15}$. $u_{15} =$ `r fitb(59, num = TRUE)`

**Q2.** A sequence is defined by $u_{n+1} = 2u_n + 1$ with $u_1 = 3$. Find $u_3$. $u_3 =$ `r fitb(15, num = TRUE)`

**Q3.** Which of these definitions lets you jump straight to the 100th term without computing any earlier terms?

(A) $u_{n+1} = 3u_n + 2$ with $u_1 = 5$  
(B) $u_n = 3n + 2$  
(C) Both of them  
(D) Neither of them

Answer: `r mcq(c("A", answer = "B", "C", "D"))`

**Q4.** An account satisfies $a_{n+1} = 1.05a_n + 100$ with $a_0 = 200$ (interest first, then a £100 deposit). Find $a_2$ in pounds, to the nearest penny. $a_2 =$ `r fitb(425.50, tol = 0.01)`

**Q5.** The rule $u_{n+1} = 3u_n - 4$ has one starting value that produces a constant sequence — its *fixed point*. Find it. $u_1 =$ `r fitb(2, num = TRUE)`
:::

`r hide("Show solutions")`

**Q1.** $59$. This is a position-to-term formula, so substitute directly: $u_{15} = 4(15) - 1 = 59$. No iteration required.

**Q2.** $15$. This is a term-to-term rule, so step through: $u_2 = 2(3) + 1 = 7$, then $u_3 = 2(7) + 1 = 15$. The trap is stopping at $7$ — that is $u_2$. The rule gives you the *next* term, so reaching $u_3$ takes two steps.

**Q3.** (B). A formula in $n$ gives any term directly: $u_{100} = 3(100) + 2 = 302$. Rule (A) looks almost identical but is a recurrence — reaching term 100 would take 99 iterations. Always check which kind of definition you have been given before "jumping ahead".

**Q4.** $425.50$. $a_1 = 1.05(200) + 100 = 310$, then $a_2 = 1.05(310) + 100 = 425.50$. Order matters: interest is applied *before* the deposit, so $1.05a_n + 100 \ne 1.05(a_n + 100)$.

**Q5.** $2$. A fixed point satisfies $u = 3u - 4$, so $2u = 4$ and $u = 2$. Verify: $3(2) - 4 = 2$ — the rule returns the value unchanged, exactly as in practice exercise 2(a).

`r unhide()`

## Sigma notation {#sigma}

### Key idea

Sigma notation compresses a sum into one expression:

$$\sum_{r=1}^{5} (2r + 1) = 3 + 5 + 7 + 9 + 11 = 35.$$

Read it as: "add up $2r + 1$ as $r$ runs from 1 to 5". The letter $r$ is a *dummy variable* — it does the counting and disappears from the answer.

Two useful splitting rules: $\sum (a_r + b_r) = \sum a_r + \sum b_r$ and $\sum k\,a_r = k \sum a_r$ for a constant $k$. Also handy: $\sum_{r=1}^{n} c = nc$ for a constant $c$.

```{r, echo=FALSE}
embed_youtube("mYoZvVz5c6s", 
              title = "Summations",
              playlist = "PLg2tfDG3Ww4umrdDAkyHGSoAfHSxImsid")
```

### Worked examples

**Example (simple).** Evaluate $\displaystyle\sum_{r=1}^{4} r^2$.

$1 + 4 + 9 + 16 = 30$.

**Example (moderate).** Evaluate $\displaystyle\sum_{r=1}^{6} (3r - 1)$.

Split: $3\sum_{r=1}^{6} r - \sum_{r=1}^{6} 1 = 3(21) - 6 = 57$.

**Example (challenging — cumulative revenue).** A start-up's revenue in month $r$ is modelled as $R_r = 200r + 1000$ (in £). Write total revenue over the first year in sigma notation and evaluate it.

$$\sum_{r=1}^{12} (200r + 1000) = 200\sum_{r=1}^{12} r + 12 \times 1000 = 200 \times 78 + 12000 = £27,600.$$

(The sum $1 + 2 + \dots + 12 = 78$ is an arithmetic series — the formula arrives in the next section.)

### Practice exercises

1. Evaluate $\displaystyle\sum_{r=1}^{5} (4r + 2)$.
2. Write the sum $2 + 4 + 8 + \dots + 256$ in sigma notation.
3. Evaluate $\displaystyle\sum_{r=1}^{20} r$.

**Common mistakes:**
- The lower limit isn't always 1: $\sum_{r=3}^{7}$ has five terms, not seven. Count terms as $\text{top} - \text{bottom} + 1$.
- $\sum_{r=1}^{n} 1 = n$, not 1 — the constant is added $n$ times.


## Arithmetic sequences and series {#arithmetic}

### Key idea

An **arithmetic sequence** changes by a constant *difference* $d$ each step: $a, \; a+d, \; a+2d, \dots$

$$\text{$n$th term:} \quad u_n = a + (n-1)d, \qquad \text{sum of $n$ terms:} \quad S_n = \frac{n}{2}\big(2a + (n-1)d\big) = \frac{n}{2}(a + \ell),$$

where $\ell$ is the last term. Arithmetic growth is *linear* growth in discrete time — the sequence version of the straight lines of Chapter \@ref(coordgeom).

### Worked examples

**Example (simple).** Find the 20th term of the sequence $5, 8, 11, \dots$

$a = 5$, $d = 3$: $u_{20} = 5 + 19 \times 3 = 62$.

**Example (moderate).** Find the sum of the first 40 terms of $7, 11, 15, \dots$

$a = 7$, $d = 4$: $S_{40} = \dfrac{40}{2}\big(14 + 39 \times 4\big) = 20 \times 170 = 3400$.

**Example (challenging — a salary scheme).** A graduate scheme pays £28,000 in year 1, rising by £1,500 each year.

(a) In which year does the salary first exceed £40,000? (b) Find the total earnings over the first 10 years.

(a) We need $28000 + (n-1) \times 1500 > 40000$, i.e. $(n-1) > 8$, so $n \ge 9$: the 9th year (salary £40,000 exactly in year 9? Check: $28000 + 8 \times 1500 = 40000$ — exactly £40,000, not exceeding it. So the salary first *exceeds* £40,000 in year 10.) Careful reading of "exceeds" versus "reaches" matters — a classic exam (and contract!) subtlety.

(b) $S_{10} = \dfrac{10}{2}\big(2 \times 28000 + 9 \times 1500\big) = 5 \times 69500 = £347,500$.

### Practice exercises

1. Find the 8th term of $4, 9, 14, \dots$
2. Find the sum of the first 25 terms of $10, 16, 22, \dots$
3. A trainee starts on £24,000, rising by £1,200 per year. Find the salary in year 6, and total earnings over the first 8 years.

**Common mistakes:**
- The $n$th term uses $(n-1)d$, not $nd$ — the first term has had *zero* increases.
- In $S_n = \tfrac{n}{2}(2a + (n-1)d)$, the $2a$ is easy to mistype as $a$. The equivalent form $\tfrac{n}{2}(\text{first} + \text{last})$ is harder to get wrong.

::: {.webex-check .webex-box}
**Self-check: arithmetic sequences and series**

**Q1.** For the sequence $9, 13, 17, 21, \dots$, give the first term and common difference. $a =$ `r fitb(9, num = TRUE)` and $d =$ `r fitb(4, num = TRUE)`

**Q2.** Find the 30th term of $6, 11, 16, \dots$

(A) $151$  
(B) $156$  
(C) $150$  
(D) $146$

Answer: `r mcq(c(answer = "A", "B", "C", "D"))`

**Q3.** Find the sum of the first 20 terms of $3, 7, 11, \dots$

(A) $790$  
(B) $820$  
(C) $860$  
(D) $1640$

Answer: `r mcq(c("A", answer = "B", "C", "D"))`

**Q4.** A salary starts at £30,000 and rises by £2,000 each year. In which year does it first *exceed* £40,000? Year: `r fitb(7, num = TRUE)`

**Q5.** An arithmetic series has first term $5$, last term $45$ and $21$ terms. Find its sum. $S_{21} =$ `r fitb(525, num = TRUE)`
:::

`r hide("Show solutions")`

**Q1.** $a = 9$, $d = 4$. The common difference comes from *subtracting* consecutive terms, $13 - 9 = 4$, and it must be the same at every step.

**Q2.** (A). $u_{30} = 6 + 29 \times 5 = 151$. The tempting (B) uses $30d$ — but the first term has had *zero* increases, so the 30th has had only 29. (C) drops the first term altogether; (D) uses $28d$.

**Q3.** (B). $S_{20} = \tfrac{20}{2}\big(2 \times 3 + 19 \times 4\big) = 10 \times 82 = 820$. The traps: (A) uses $a$ where $2a$ belongs; (C) uses $20d$ instead of $19d$; (D) forgets the $\tfrac{n}{2}$ entirely. Cross-check with the harder-to-fumble form $\tfrac{n}{2}(\text{first} + \text{last})$: the last term is $3 + 19 \times 4 = 79$, and $10(3 + 79) = 820$.

**Q4.** Year $7$. Year $n$ pays $30000 + (n-1) \times 2000$. Year 6 pays exactly £40,000 — that *reaches* the figure but does not exceed it. The first year strictly above is year 7, at £42,000. "Exceeds" versus "reaches" matters — in exams and in contracts.

**Q5.** $525$. With the first and last terms known, $S_n = \tfrac{n}{2}(a + \ell) = \tfrac{21}{2}(5 + 45) = 21 \times 25 = 525$.

`r unhide()`

## Geometric sequences and series {#geometric}

### Key idea

A **geometric sequence** changes by a constant *ratio* $r$ each step: $a, \; ar, \; ar^2, \dots$

$$\text{$n$th term:} \quad u_n = ar^{n-1}, \qquad \text{sum of $n$ terms:} \quad S_n = \frac{a(1 - r^n)}{1 - r} \quad (r \ne 1).$$

When $|r| < 1$ the terms shrink towards zero and the infinite sum settles at

$$S_\infty = \frac{a}{1 - r}.$$

Geometric growth is *percentage* growth: multiplying by $1.05$ each year is 5% compound growth; multiplying by $0.85$ is 15% decay.

### Worked examples

**Example (simple).** Find the 10th term of $3, 6, 12, \dots$

$a = 3$, $r = 2$: $u_{10} = 3 \times 2^9 = 1536$.

**Example (moderate).** Find the sum of the first 8 terms of $100, 90, 81, \dots$ to 2 d.p.

$a = 100$, $r = 0.9$: $S_8 = \dfrac{100(1 - 0.9^8)}{1 - 0.9} = 1000(1 - 0.43046721) = 569.53.$

**Example (challenging — the long tail of a promotion).** A viral marketing campaign attracts 5,000 new customers in its first month. Each month afterwards, the effect fades: new sign-ups are 80% of the previous month's. If the pattern continued forever, how many customers would the campaign bring in total?

Monthly sign-ups form a geometric sequence with $a = 5000$, $r = 0.8$. Since $|r| < 1$:

$$S_\infty = \frac{5000}{1 - 0.8} = 25,000.$$

The campaign is worth 25,000 customers *in total* — a finite answer to an infinite sum, and exactly the calculation behind "customer lifetime value" and the valuation of perpetuities in finance. Also useful the other way: after 12 months the campaign has delivered $S_{12} = \tfrac{5000(1 - 0.8^{12})}{0.2} \approx 24,313$ customers — over 97% of everything it will ever deliver.

```{r geometric-decay, echo=FALSE, fig.cap="Monthly sign-ups (bars) fade geometrically while the cumulative total (line) approaches its limit of 25,000.", fig.height=4}
n <- 1:15
signups <- 5000 * 0.8^(n - 1)
cumulative <- cumsum(signups)
plot(n, cumulative, type = "b", pch = 16, col = "steelblue", lwd = 2,
     xlab = "Month", ylab = "Customers", ylim = c(0, 26000))
segments(n, 0, n, signups, lwd = 6, col = "grey60")
abline(h = 25000, lty = 2, col = "firebrick")
text(13.5, 23800, expression(S[infinity] == 25000), col = "firebrick")
```

### Practice exercises

1. Find the 7th term of $2, 6, 18, \dots$
2. Find the sum of the first 10 terms of $5, 10, 20, \dots$
3. Find $S_\infty$ for the series $12 + 9 + 6.75 + \dots$
4. A subscription box gains 800 sign-ups in month 1, and each subsequent month brings 70% as many as the month before. Estimate the total sign-ups the launch will ever generate.

**Common mistakes:**

- The ratio is found by *dividing* consecutive terms, not subtracting: for $100, 90, 81$, $r = 0.9$ (and the difference isn't even constant).

- $S_\infty$ exists **only** when $|r| < 1$. For $r = 1.05$ the sum grows without bound — quoting $\tfrac{a}{1-r}$ (a negative number!) is a sure sign something's gone wrong.

- $u_n = ar^{n-1}$: the 10th term has $r^9$, not $r^{10}$.

```{r, echo=FALSE}
embed_desmos("hbawd52hhk")
```

**Why this matters:** Nearly every formula in a first finance course — compound interest, annuities, mortgages, bond prices, perpetuities, discounted cash flow — is a dressed-up geometric series. Recognise the pattern "same percentage change every period" and you already know the mathematics.

## Binomial expansion {#binomial}

### Key idea

For a positive integer $n$,

$$(a + b)^n = \sum_{k=0}^{n} \binom{n}{k} a^{n-k} b^k, \qquad \binom{n}{k} = \frac{n!}{k!(n-k)!},$$

where the binomial coefficients $\binom{n}{k}$ (also written $^n\mathrm{C}_k$) can be read off Pascal's triangle. Two practical uses: expanding brackets like $(2 + 3x)^4$ quickly, and **approximating** powers — if $x$ is small, the first few terms of $(1 + x)^n$ are an excellent estimate because later terms involve tiny powers of $x$.

### Worked examples

**Example (simple).** Expand $(1 + x)^4$.

Row 4 of Pascal's triangle is $1, 4, 6, 4, 1$:

$$(1+x)^4 = 1 + 4x + 6x^2 + 4x^3 + x^4.$$

**Example (moderate).** Expand $(2 + 3x)^4$.

$$\binom{4}{0}2^4 + \binom{4}{1}2^3(3x) + \binom{4}{2}2^2(3x)^2 + \binom{4}{3}2(3x)^3 + \binom{4}{4}(3x)^4$$

$$= 16 + 96x + 216x^2 + 216x^3 + 81x^4.$$

Notice every term uses a power of 2 *and* a power of $3x$, and the powers always total 4.

**Example (challenging — a quick compound-interest estimate).** Use the first three terms of a binomial expansion to estimate $1.02^{10}$, and compare with the true value. What does this say about 2% growth over 10 years?

Write $1.02^{10} = (1 + 0.02)^{10}$:

$$(1 + 0.02)^{10} \approx 1 + 10(0.02) + \binom{10}{2}(0.02)^2 = 1 + 0.2 + 45 \times 0.0004 = 1.218.$$

The true value is $1.21899\ldots$ — the three-term estimate is off by less than 0.1%. Interpretation: 2% annual growth for 10 years gives about 21.8% total growth, *not* 20%. The first term of the correction, $+0.018$, is precisely the "interest on interest" that simple-interest thinking misses.

### Practice exercises

1. Expand $(1 + 2x)^3$.
2. Find the coefficient of $x^2$ in the expansion of $(3 + x)^5$.
3. Use the first three terms of $(1 - 0.02)^5$ to estimate $0.98^5$, to 3 d.p.

**Common mistakes:**

- In $(2 + 3x)^4$, the powers of 2 are as important as the powers of $3x$
— forgetting to raise the first term is the classic slip.
- $(3x)^2 = 9x^2$, not $3x^2$: the coefficient gets squared too.
- Approximations need $x$ *small*. Truncating $(1 + 0.5)^{10}$ after three terms would be badly wrong.

```{r, echo=FALSE}
embed_youtube("x0mZ9HaltU4", 
              title = "Binomial expansions",
              playlist = "PLg2tfDG3Ww4tDiNfyq94pE0BI3wiCVTps")
```
::: {.webex-check .webex-box}
**Self-check: binomial expansion**

**Q1.** Evaluate $\dbinom{6}{2}$. Answer: `r fitb(15, num = TRUE)`

**Q2.** In the expansion of $(1 + 2x)^3$, find the coefficient of $x^2$. Coefficient: `r fitb(12, num = TRUE)`

**Q3.** Find the coefficient of $x^3$ in the expansion of $(2 + x)^5$.

(A) $10$  
(B) $20$  
(C) $40$  
(D) $80$

Answer: `r mcq(c("A", "B", answer = "C", "D"))`

**Q4.** Use the first three terms of the binomial expansion to estimate $1.03^8$, to 3 decimal places. Estimate: `r fitb(1.265, tol = 0.001)`

**Q5.** Truncating $(1 + x)^{10}$ after three terms gives a good approximation only when $x$ is small. For which of these values of $x$ would the estimate be *poor*? `r mcq(c("0.01", "0.05", answer = "0.5"))`
:::

`r hide("Show solutions")`

**Q1.** $15$. $\dbinom{6}{2} = \dfrac{6 \times 5}{2 \times 1} = 15$ — or read it from row 6 of Pascal's triangle: $1, 6, 15, 20, 15, 6, 1$.

**Q2.** $12$. The $x^2$ term is $\dbinom{3}{2}(2x)^2 = 3 \times 4x^2 = 12x^2$. The trap answer $6$ comes from writing $(2x)^2 = 2x^2$ — the coefficient gets squared too.

**Q3.** (C). The $x^3$ term is $\dbinom{5}{3} \times 2^2 \times x^3 = 10 \times 4 = 40x^3$. (A) forgets the powers of 2 altogether; (D) uses $2^3$ — but the powers of $2$ and of $x$ must total 5, so $x^3$ pairs with $2^2$. Quick sanity check: with $x = 0$ the expansion must give $2^5 = 32$, which only works if the powers of 2 are kept throughout.

**Q4.** $1.265$. $(1 + 0.03)^8 \approx 1 + 8(0.03) + \dbinom{8}{2}(0.03)^2 = 1 + 0.24 + 28 \times 0.0009 = 1.2652$. The true value is $1.26677\ldots$, so three terms already land within $0.002$ — 3% growth over 8 years is about 26.7%, not 24%.

**Q5.** $0.5$. The approximation relies on higher powers of $x$ being negligible, and $0.5^3 = 0.125$ is anything but. For $x = 0.01$ or $0.05$ the neglected tail really is tiny.

`r unhide()`

## Modelling with sequences and series {#seqmodelling}

### Key idea

Discrete-time models describe quantities updated period by period:

- **Constant amount per period** → arithmetic sequence (linear growth): straight-line depreciation, fixed increments.
- **Constant percentage per period** → geometric sequence (exponential growth or decay): compound interest, reducing-balance depreciation, fading demand.
- **Percentage change *plus* a regular payment** → a recurrence of the form $u_{n+1} = ru_n + c$: savings plans, loan repayment, inventory with restocking. Geometric series turn these recurrences into closed formulae.

Chapter \@ref(explog) develops the continuous-time versions of these models and (crucially) the logarithms that solve "how long until…?" questions exactly.

### Worked examples

**Example (simple).** A machine bought for £20,000 loses 15% of its value each year. Write a formula for its value after $n$ years, and find the value after 4 years.

Multiplying by $0.85$ each year: $V_n = 20000 \times 0.85^n$. After 4 years, $V_4 = 20000 \times 0.52200625 \approx £10,440$.

Compare straight-line depreciation (arithmetic): reducing-balance loses *more* early and *less* later — often more realistic for equipment.

**Example (moderate).** £2,000 is invested at 4% per year compound. After how many complete years does the investment first exceed £3,000?

Balance after $n$ years: $2000 \times 1.04^n$. We need $1.04^n > 1.5$. Trying values: $1.04^{10} \approx 1.480$ (not yet); $1.04^{11} \approx 1.539$ (yes). So 11 years.

Trial-and-improvement works, but it's clumsy — in Chapter \@ref(explog), logarithms will solve $1.04^n = 1.5$ in one line.

**Example (challenging — paying off a loan).** A firm borrows £10,000 at 1% interest per month and repays £500 at the end of each month. Let $b_n$ be the balance after $n$ payments.

(a) Write down a recurrence for $b_n$. (b) Show, using a geometric series, that $b_n = 50000 - 40000 \times 1.01^n$. (c) Roughly how many months does repayment take?

(a) Interest, then payment: $b_{n+1} = 1.01\,b_n - 500$, with $b_0 = 10000$.

(b) Unrolling the recurrence: each month the debt is multiplied by $1.01$ and every past £500 payment has been compounding *against* the debt since it was made:

$$b_n = 10000 \times 1.01^n - 500\left(1.01^{n-1} + 1.01^{n-2} + \dots + 1\right) = 10000 \times 1.01^n - 500 \cdot \frac{1.01^n - 1}{0.01}.$$

The geometric-series formula collapses the bracket, giving $b_n = 10000 \times 1.01^n - 50000(1.01^n - 1) = 50000 - 40000 \times 1.01^n$.

(c) The loan clears when $b_n = 0$: $1.01^n = 1.25$. Trial gives $n \approx 22.4$, so the firm makes 22 full payments of £500 and a smaller 23rd payment. Every mortgage calculator on the internet is running exactly this algebra.

```{r loan-balance, echo=FALSE, fig.cap="Loan balance under the recurrence b(n+1) = 1.01 b(n) - 500: slow progress early (interest bites), faster later.", fig.height=4}
b <- numeric(24); b[1] <- 10000
for (i in 1:23) b[i + 1] <- max(1.01 * b[i] - 500, 0)
plot(0:23, b, type = "b", pch = 16, col = "steelblue", lwd = 2,
     xlab = "Payments made", ylab = "Balance (£)")
abline(h = 0, col = "grey70")
```

### Practice exercises

1. A car bought for £18,000 loses 12% of its value each year. Find its value after 5 years, to the nearest pound.
2. £1,500 is invested at 3.5% per year compound. Find its value after 6 years, to the nearest pound.
3. A £5,000 loan charges 2% interest per month, with £300 repaid at the end of each month. Write the recurrence for the balance and compute the balance after two payments.

**Common mistakes:**

- A 15% *decrease* means multiplying by $0.85$ — not by $0.15$, and not subtracting $0.15$.
- Percentage changes compound: two years of 10% growth gives a factor of $1.1^2 = 1.21$ (21%), not 20%.
- In "interest then payment" recurrences, order matters: $1.01b_n - 500 \ne 1.01(b_n - 500)$. Read the account's rules carefully.

**Why this matters:** These recurrences *are* financial mathematics: annuities, mortgages, sinking funds and net present value are all variations on $u_{n+1} = ru_n + c$. The same structure models inventory (stock decays, deliveries top it up) and even customer retention (churn plus acquisition). Learn one recurrence, unlock a dozen models.

## Chapter summary

You should now be able to:

- work with sequences given by formulae or by recurrence relations, and iterate term by term;
- read, write and evaluate sums in sigma notation;
- find $n$th terms and sums of arithmetic sequences, and use them for constant-increment models;
- find $n$th terms, finite sums and (for $|r| < 1$) infinite sums of geometric sequences;
- expand $(a + bx)^n$ for positive integer $n$ and use truncated expansions to approximate;
- build and solve discrete-time models for depreciation, compound growth, savings and loans.

The unfinished business — solving equations like $1.04^n = 1.5$ exactly — is precisely where Chapter \@ref(explog) picks up.

## End-of-chapter diagnostic {#sequences-diagnostic}

This diagnostic samples the whole chapter — treat it as a compass, not a verdict: it tells you where to head next, nothing more.

## End of chapter quiz

::: {.webex-check .webex-box}
**Q1.** A sequence is defined by $u_{n+1} = 5u_n - 6$ with $u_1 = 2$. Find $u_3$. $u_3 =$ `r fitb(14, num = TRUE)`

**Q2.** Evaluate $\displaystyle\sum_{r=1}^{10} (2r + 3)$. Answer: `r fitb(140, num = TRUE)`

**Q3.** Find the 12th term of $8, 15, 22, \dots$ $u_{12} =$ `r fitb(85, num = TRUE)`

**Q4.** Find the sum of the first 30 terms of $2, 5, 8, \dots$

(A) $1335$  
(B) $1365$  
(C) $1410$  
(D) $2730$

Answer: `r mcq(c("A", answer = "B", "C", "D"))`

**Q5.** Find the 8th term of $5, 10, 20, \dots$ $u_8 =$ `r fitb(640, num = TRUE)`

**Q6.** Find the sum of the first 6 terms of $4, 12, 36, \dots$

(A) $1456$  
(B) $2912$  
(C) $484$  
(D) $972$

Answer: `r mcq(c(answer = "A", "B", "C", "D"))`

**Q7.** For the series $20 + 15 + 11.25 + \dots$, find $S_\infty$. $S_\infty =$ `r fitb(80, num = TRUE)`. Does $S_\infty$ exist for the series $6 + 9 + 13.5 + \dots$? `r mcq(c("yes", answer = "no"))`

**Q8.** Find the coefficient of $x^2$ in the expansion of $(3 + 2x)^4$. Coefficient: `r fitb(216, num = TRUE)`

**Q9.** Use the first three terms of the binomial expansion to estimate $0.99^{10}$, to 4 decimal places. Estimate: `r fitb(0.9045, tol = 0.0005)`

**Q10.** A machine bought for £15,000 loses 20% of its value each year. Find its value after 3 years, in pounds. Value: `r fitb(7680, num = TRUE)`

**Q11.** A loan charges 1.5% interest per month, and £400 is repaid at the end of each month. Which recurrence describes the balance?

(A) $b_{n+1} = 1.5\,b_n - 400$  
(B) $b_{n+1} = 1.015(b_n - 400)$  
(C) $b_{n+1} = 1.015\,b_n - 400$  
(D) $b_{n+1} = 0.015\,b_n - 400$

Answer: `r mcq(c("A", "B", answer = "C", "D"))`

**Q12.** £3,000 is invested at 5% per year compound. After how many complete years does the investment first exceed £4,000? Years: `r fitb(6, num = TRUE)`
:::

`r hide("Show solutions")`

**Q1.** $14$. Step through: $u_2 = 5(2) - 6 = 4$, then $u_3 = 5(4) - 6 = 14$. Stopping at $4$ gives $u_2$, not $u_3$ — recurrences must be iterated. *Shaky? Revisit Section \@ref(recurrence).*

**Q2.** $140$. Split the sum: $2\sum_{r=1}^{10} r + \sum_{r=1}^{10} 3 = 2(55) + 30 = 140$. The constant is added *ten times* — $\sum_{r=1}^{10} 3 = 30$, not $3$. *Shaky? Revisit Section \@ref(sigma).*

**Q3.** $85$. $a = 8$, $d = 7$: $u_{12} = 8 + 11 \times 7 = 85$. Eleven increases, not twelve — the first term has had none. *Shaky? Revisit Section \@ref(arithmetic).*

**Q4.** (B). $S_{30} = \tfrac{30}{2}\big(2 \times 2 + 29 \times 3\big) = 15 \times 91 = 1365$. (A) uses $a$ where $2a$ belongs, (C) uses $30d$, (D) drops the $\tfrac{n}{2}$. Cross-check: the last term is $2 + 29 \times 3 = 89$, and $\tfrac{30}{2}(2 + 89) = 1365$. *Shaky? Revisit Section \@ref(arithmetic).*

**Q5.** $640$. $a = 5$, $r = 2$: $u_8 = 5 \times 2^7 = 640$. The 8th term uses $r^7$, not $r^8$. *Shaky? Revisit Section \@ref(geometric).*

**Q6.** (A). $S_6 = \dfrac{4(3^6 - 1)}{3 - 1} = \dfrac{4 \times 728}{2} = 1456$. (B) forgets to divide by $r - 1 = 2$; (C) uses $3^5$ in the formula; (D) is the 6th *term*, $4 \times 3^5 = 972$, not the sum. *Shaky? Revisit Section \@ref(geometric).*

**Q7.** $80$, and **no**. First series: $r = \tfrac{15}{20} = 0.75$, so $S_\infty = \dfrac{20}{1 - 0.75} = 80$. Second series: $r = 1.5$, and $S_\infty$ exists only when $|r| < 1$ — the terms grow, so the sum grows without bound. Quoting $\tfrac{a}{1-r}$ there would give a negative "sum": a sure sign something has gone wrong. *Shaky? Revisit Section \@ref(geometric).*

**Q8.** $216$. The $x^2$ term is $\dbinom{4}{2} \times 3^2 \times (2x)^2 = 6 \times 9 \times 4x^2 = 216x^2$. The powers of $3$ matter as much as the powers of $2x$, and $(2x)^2 = 4x^2$ — the coefficient gets squared. *Shaky? Revisit Section \@ref(binomial).*

**Q9.** $0.9045$. $(1 - 0.01)^{10} \approx 1 + 10(-0.01) + \dbinom{10}{2}(-0.01)^2 = 1 - 0.1 + 0.0045 = 0.9045$. Watch the signs: the squared term comes back *positive*. (True value $0.90438\ldots$) *Shaky? Revisit Section \@ref(binomial).*

**Q10.** $7680$. Losing 20% means multiplying by $0.8$: $15000 \times 0.8^3 = 7680$. Not multiplying by $0.2$, and not subtracting 20% three times — percentage changes compound. *Shaky? Revisit Section \@ref(seqmodelling).*

**Q11.** (C). Interest multiplies the *whole balance* by $1.015$, then £400 comes off. (A) confuses a 1.5% rate with a factor of $1.5$; (B) applies the payment before the interest — order matters, since $1.015b_n - 400 \ne 1.015(b_n - 400)$; (D) keeps the interest but throws away the balance itself. *Shaky? Revisit Section \@ref(seqmodelling).*

**Q12.** $6$. We need $3000 \times 1.05^n > 4000$, i.e. $1.05^n > 1.333\ldots$ Trying values: $1.05^5 \approx 1.276$ (not yet), $1.05^6 \approx 1.340$ (yes). Chapter \@ref(explog) will replace the trial-and-error with a one-line logarithm. *Shaky? Revisit Section \@ref(seqmodelling).*

`r unhide()`

**Rule of thumb:** if two or three questions from the same section went wrong, rework that section before starting the next chapter — the geometric-series and modelling material in particular feeds directly into Chapter \@ref(explog).

## Answers to practice exercises {#sequences-answers}

**Sequences and recurrence relations:** 1. $3, 8, 13$; $u_{12} = 58$.  2. (a) $2, 2, 2$ — the value 2 is a fixed point of the rule; (b) $5, 11, 29$.  3. $a_1 = 560$, $a_2 = 621.20$, $a_3 = 683.62$.

**Sigma notation:** 1. $4(1+2+3+4+5) + 10 = 70$.  2. $\displaystyle\sum_{r=1}^{8} 2^r$.  3. $210$.

**Arithmetic:** 1. $u_8 = 4 + 7 \times 5 = 39$.  2. $S_{25} = \tfrac{25}{2}(20 + 24 \times 6) = 2050$.  3. Year 6: £30,000; total: $\tfrac{8}{2}(48000 + 7 \times 1200) = £225,600$.

**Geometric:** 1. $2 \times 3^6 = 1458$.  2. $5(2^{10} - 1) = 5115$.  3. $r = 0.75$, $S_\infty = \tfrac{12}{0.25} = 48$.  4. $\tfrac{800}{0.3} \approx 2667$ sign-ups.

**Binomial:** 1. $1 + 6x + 12x^2 + 8x^3$.  2. $\binom{5}{2}3^3 = 270$.  3. $1 - 0.1 + 0.004 = 0.904$ (true value $0.9039\ldots$).

**Modelling:** 1. $18000 \times 0.88^5 \approx £9,499$.  2. $1500 \times 1.035^6 \approx £1,844$.  3. $b_{n+1} = 1.02b_n - 300$, $b_0 = 5000$; $b_1 = 4800$, $b_2 = £4,596$.
