# Integration {#integration}

Integration runs differentiation in reverse — and that reversal is astonishingly useful. Chapter \@ref(differentiation) turned totals into rates: cost into marginal cost, sales into sales growth. Integration turns rates back into totals: marginal cost back into cost, a sales *rate* into total units sold, a flow into an accumulated amount. It also computes areas, and — the finale — solves the differential equations you learned to construct in Section \@ref(constructing-des).

Like the last chapter, this one is long and example-heavy: the techniques (substitution, parts, partial fractions) take practice, and the interpretation questions matter as much as the mechanics.

**Why this matters:** "Total from rate" is half of business analytics: total revenue from a revenue rate, total demand from a demand intensity, accumulated interest from an interest rate. Areas under curves have direct economic meanings — consumer surplus is one you'll compute in this chapter — and separable differential equations give you exponential and saturating growth *derived*, not assumed. Integration also underpins all of probability (areas under density curves), which dominates your statistics modules.

## Integration as the reverse of differentiation {#antiderivatives}

### Key idea

An **antiderivative** (or indefinite integral) of $f(x)$ is any function $F(x)$ whose derivative is $f(x)$. We write

$$\int f(x)\,dx = F(x) + C,$$

where $C$ is the **constant of integration** — needed because constants vanish under differentiation, so infinitely many functions share the same derivative. The **Fundamental Theorem of Calculus** makes the reversal precise: differentiation and integration undo each other, and (as we'll use in Section \@ref(definite-integrals)) areas under $f$ can be computed from any antiderivative $F$.

The constant $C$ is not a nuisance — in applications it carries real information, pinned down by an **initial condition**.

### Worked examples

**Example (simple).** Find $\displaystyle\int (6x^2 - 4x + 1)\,dx$.

Reverse the power rule term by term (raise the power, divide by the new power):

$$2x^3 - 2x^2 + x + C.$$

Check by differentiating — you should recover exactly $6x^2 - 4x + 1$. *Always check*: differentiation is easier than integration, so the check is cheap insurance.

**Example (moderate).** Given $\dfrac{dy}{dx} = 6\sqrt{x}$ and $y = 30$ when $x = 4$, find $y$.

Rewrite and integrate: $\displaystyle\int 6x^{1/2}\,dx = 4x^{3/2} + C$.

Apply the condition: $30 = 4(4)^{3/2} + C = 32 + C$, so $C = -2$ and $y = 4x^{3/2} - 2$.

**Example (challenging — recovering total cost from marginal cost).** A firm's marginal cost is $C'(q) = 15 - 1.2q + 0.03q^2$ (the function from Section \@ref(standard-derivatives)), and its fixed costs are £200. Find the total cost function.

Integrate:

$$C(q) = \int (15 - 1.2q + 0.03q^2)\,dq = 15q - 0.6q^2 + 0.01q^3 + C.$$

Fixed costs are the costs at zero output: $C(0) = 200$, so $C = 200$ and

$$C(q) = 200 + 15q - 0.6q^2 + 0.01q^3.$$

We have reconstructed Chapter \@ref(differentiation)'s cost function exactly — and discovered something worth framing: **the constant of integration is the fixed cost**. Abstract constant, concrete meaning.

### Practice exercises

1. Find $F(x)$ given $F'(x) = 8x^3 - 3x^2 + 2$.
2. Given $\dfrac{dy}{dx} = \dfrac{3}{\sqrt{x}}$ and $y(9) = 20$, find $y$.
3. Marginal revenue is $R'(q) = 50 - 0.4q$ and $R(0) = 0$ (no sales, no revenue). Find the revenue function.

**Common mistakes:**

- Forgetting $+C$ on indefinite integrals. It costs marks at A-level and *meaning* in applications (it's the fixed cost, the starting balance, the initial population…).
- Reversing the power rule the wrong way round: integrate $x^n$ by *raising* the power to $n+1$ and dividing by $n+1$ — not multiplying.
- Not checking by differentiating. Thirty seconds, catches nearly everything.

```{r, echo=FALSE}
embed_youtube("h5kWi4IiYiE", 
              title = "The Fundamental Theorem of Calculus")
```

::: {.webex-check .webex-box}
**Self-check: antiderivatives and initial conditions**

**Q1.** $\displaystyle\int 10x^4\,dx = kx^5 + C$. $k =$ `r fitb(2, num = TRUE)`

**Q2.** Which of the following equals $\displaystyle\int (9x^2 - 4x + 3)\,dx$?

(A) $9x^3 - 4x^2 + 3x + C$  
(B) $18x - 4$  
(C) $3x^3 - 2x^2 + 3x + C$  
(D) $3x^3 - 2x^2 + 3x$  

Answer: `r mcq(c("A", "B", answer = "C", "D"))`

**Q3.** Given $\dfrac{dy}{dx} = 4x - 5$ and $y(2) = 1$, find the constant of integration. $C =$ `r fitb(3, num = TRUE)`

**Q4.** Given $\dfrac{dy}{dx} = 3\sqrt{x}$ and $y(4) = 20$: $C =$ `r fitb(4, num = TRUE)` and $y(1) =$ `r fitb(6, num = TRUE)`

**Q5.** A firm's marginal cost is $C'(q) = 20 - 0.5q$ and its fixed costs are £150. Find the total cost of producing 10 units, in £: `r fitb(325, num = TRUE)`. In the total cost function, the constant of integration represents the `r mcq(c("marginal cost", answer = "fixed cost", "variable cost", "profit"))`.
:::

`r hide("Show solutions")`
**Q1.** $k = 2$: raise the power to 5, then *divide* by 5, so $\tfrac{10}{5} = 2$. Multiplying by the new power gives 50 — the power rule run the wrong way round.

**Q2.** (C). Integrate term by term: $3x^3 - 2x^2 + 3x + C$. (B) is the *derivative* — integration goes the other way. (A) raises each power without dividing by it. (D) is correct but missing $+C$: incomplete, and in applications the constant carries real meaning.

**Q3.** $y = 2x^2 - 5x + C$; then $y(2) = 8 - 10 + C = 1$, so $C = 3$.

**Q4.** $\displaystyle\int 3x^{1/2}\,dx = 2x^{3/2} + C$; $y(4) = 2(8) + C = 20$ gives $C = 4$, so $y(1) = 2 + 4 = 6$. Quick check: $\tfrac{d}{dx}\big(2x^{3/2}\big) = 3x^{1/2}$ ✓.

**Q5.** $C(q) = 150 + 20q - 0.25q^2$, so $C(10) = 150 + 200 - 25 = 325$. The condition $C(0) = 150$ pins the constant of integration: it *is* the fixed cost — abstract constant, concrete meaning.
`r unhide()`

## Standard integrals {#standard-integrals}

### Key idea

The standard results, each the reverse of a Chapter \@ref(differentiation) derivative:

$$\int x^n\,dx = \frac{x^{n+1}}{n+1} + C \;\; (n \ne -1), \qquad \int e^x\,dx = e^x + C, \qquad \int \frac{1}{x}\,dx = \ln|x| + C,$$

together with the structural rules: constants multiply through, sums integrate term by term. Note the exception in the power rule — $n = -1$ would mean dividing by zero, and that missing case is precisely where $\ln|x|$ steps in. As in differentiation: **rewrite first** (surds and reciprocals as powers, quotients simplified), integrate second.

### Worked examples

**Example (simple).** Find $\displaystyle\int \left(x^4 - 2x + 7\right) dx$.

$$\frac{x^5}{5} - x^2 + 7x + C.$$

**Example (moderate).** Find $\displaystyle\int \left(6\sqrt{x} + \frac{3}{x^2}\right) dx$.

Rewrite: $6x^{1/2} + 3x^{-2}$. Integrate:

$$6 \cdot \frac{x^{3/2}}{3/2} + 3 \cdot \frac{x^{-1}}{-1} + C = 4x^{3/2} - \frac{3}{x} + C.$$

**Example (moderate).** Find $\displaystyle\int \frac{2x^3 - 5x}{x}\,dx$.

Simplify first: $\dfrac{2x^3 - 5x}{x} = 2x^2 - 5$, so the integral is $\dfrac{2x^3}{3} - 5x + C$. (The simplify-first habit from Chapter \@ref(differentiation) pays double here — there is no "quotient rule" for integration to rescue you.)

**Example (challenging).** Find $\displaystyle\int \left(3e^x + \frac{2}{x} - \frac{4}{x^3}\right) dx$.

Treat each term with its own rule — and note which is which: $\tfrac{2}{x}$ is the special $\ln$ case; $\tfrac{4}{x^3} = 4x^{-3}$ is an ordinary power:

$$3e^x + 2\ln|x| + \frac{2}{x^2} + C.$$

(Check the last term: $\tfrac{d}{dx}\big(2x^{-2}\big) = -4x^{-3}$ ✓.)

### Practice exercises

1. Find $\displaystyle\int (4x^3 - 6x^2 + 2)\,dx$.
2. Find $\displaystyle\int \left(\frac{6}{\sqrt{x}} - \frac{4}{x^3}\right) dx$.
3. Find $\displaystyle\int \frac{x^3 + 4x}{x^2}\,dx$ (simplify first).
4. Find $\displaystyle\int \left(5e^x - \frac{3}{x}\right) dx$.

**Common mistakes:**

- Applying the power rule to $\tfrac{1}{x}$: $\int x^{-1}dx$ is $\ln|x| + C$, never $\tfrac{x^0}{0}$.
- Sign and power slips on negative powers: $\int x^{-3}dx = \tfrac{x^{-2}}{-2} = -\tfrac{1}{2x^2}$.
- Trying to integrate a product or quotient term by term. $\int f g \,dx \ne \int f\,dx \int g\,dx$ — products need Sections \@ref(substitution)–\@ref(byparts).

::: {.webex-check .webex-box}
**Self-check: standard integrals (quick-fire)**

**Q1.** $\displaystyle\int x^6\,dx = \dfrac{x^7}{n} + C$. $n =$ `r fitb(7, num = TRUE)`

**Q2.** $\displaystyle\int 5e^x\,dx = k\,e^x + C$. $k =$ `r fitb(5, num = TRUE)`

**Q3.** Which equals $\displaystyle\int \dfrac{2}{x}\,dx$?

(A) $-\dfrac{2}{x^2} + C$  
(B) $2\ln|x| + C$  
(C) $\dfrac{2x^0}{0} + C$  
(D) $\ln|2x| + C$  

Answer: `r mcq(c("A", answer = "B", "C", "D"))`

**Q4.** Rewrite first: $\displaystyle\int 9\sqrt{x}\,dx = k\,x^{3/2} + C$. $k =$ `r fitb(6, num = TRUE)`

**Q5.** Rewrite first: $\displaystyle\int \dfrac{4}{x^3}\,dx = \dfrac{k}{x^2} + C$. $k =$ `r fitb(-2, num = TRUE)`

**Q6.** Which equals $\displaystyle\int \dfrac{1}{\sqrt{x}}\,dx$?

(A) $\sqrt{x} + C$  
(B) $\ln\sqrt{x} + C$  
(C) $-\tfrac{1}{2}x^{-3/2} + C$  
(D) $2\sqrt{x} + C$  

Answer: `r mcq(c("A", "B", "C", answer = "D"))`

**Q7.** Simplify first: $\displaystyle\int \dfrac{x^3 + 6x}{x}\,dx = \dfrac{x^3}{a} + bx + C$. $a =$ `r fitb(3, num = TRUE)` and $b =$ `r fitb(6, num = TRUE)`

**Q8.** Which equals $\displaystyle\int \left(3e^x - \dfrac{2}{x^2}\right) dx$?

(A) $3e^x + \dfrac{2}{x} + C$  
(B) $3e^x - \dfrac{2}{x} + C$  
(C) $3e^x - 2\ln\big(x^2\big) + C$  
(D) $3e^x + \dfrac{2}{3x^3} + C$  

Answer: `r mcq(c(answer = "A", "B", "C", "D"))`

**Q9.** Simplify first: $\displaystyle\int 2x \cdot x^3\,dx = k\,x^5 + C$. $k$ (as a decimal) $=$ `r fitb(0.4, tol = 0.01)`

**Q10.** Which equals $\displaystyle\int \left(\dfrac{4}{x} + \dfrac{x}{4}\right) dx$?

(A) $4\ln|x| + \dfrac{x^2}{4} + C$  
(B) $-\dfrac{4}{x^2} + \dfrac{1}{4} + C$  
(C) $4\ln|x| + \dfrac{x^2}{8} + C$  
(D) $\ln|4x| + \dfrac{x^2}{8} + C$  

Answer: `r mcq(c("A", "B", answer = "C", "D"))`
:::

`r hide("Show solutions")`
**Q1.** $n = 7$: raise the power, divide by the new power.

**Q2.** $k = 5$: $e^x$ is its own antiderivative; the constant multiplies through.

**Q3.** (B). This is the power rule's missing case: (C) is what the blind power rule produces — division by zero — and exactly where $\ln|x|$ steps in. (A) is the derivative. (D) looks close, but $\ln|2x| = \ln 2 + \ln|x|$ has derivative $\tfrac{1}{x}$, not $\tfrac{2}{x}$: the 2 must multiply the log, not sit inside it.

**Q4.** $9x^{1/2}$ integrates to $9 \cdot \tfrac{x^{3/2}}{3/2} = 6x^{3/2}$: dividing by $\tfrac32$ means multiplying by $\tfrac23$.

**Q5.** $4x^{-3}$ integrates to $\tfrac{4x^{-2}}{-2} = -\tfrac{2}{x^2}$, so $k = -2$. Negative powers are sign-slip territory: check by differentiating.

**Q6.** (D). $x^{-1/2}$ integrates to $\tfrac{x^{1/2}}{1/2} = 2\sqrt{x}$. (C) is the derivative; (B) misreads "reciprocal" as "logarithm" — only $x^{-1}$ integrates to a log; (A) forgot to divide by $\tfrac12$.

**Q7.** Simplify to $x^2 + 6$ first, giving $\tfrac{x^3}{3} + 6x + C$: $a = 3$, $b = 6$. There is no quotient rule for integration — simplify or nothing.

**Q8.** (A). The second term is an ordinary power: $\displaystyle\int -2x^{-2}\,dx = -2 \cdot \tfrac{x^{-1}}{-1} = +\tfrac{2}{x}$. (C) treats $\tfrac{1}{x^2}$ as the log case — only $\tfrac1x$ qualifies; (D) moves the power the wrong way; (B) is the sign slip.

**Q9.** Simplify the product to $2x^4$ first: $\tfrac{2x^5}{5}$, so $k = 0.4$. Never integrate a product factor by factor — $\int fg\,dx \ne \int f\,dx \int g\,dx$.

**Q10.** (C). $\displaystyle\int \tfrac{x}{4}\,dx = \tfrac{x^2}{8}$ — the tempting $\tfrac{x^2}{4}$ in (A) forgets the extra division by 2. (D)'s $\ln|4x|$ differentiates to $\tfrac1x$, not $\tfrac4x$; (B) is the derivative.
`r unhide()`

## Definite integrals and area {#definite-integrals}

### Key idea

The **definite integral** evaluates the accumulated change of an antiderivative between limits:

$$\int_a^b f(x)\,dx = \Big[F(x)\Big]_a^b = F(b) - F(a),$$

and equals the **area between the curve and the $x$-axis** from $a$ to $b$ — with one crucial caveat: regions *below* the axis count as **negative**. For genuine areas, find where the curve crosses the axis and treat each piece separately. The area **between two curves** is

$$\int_a^b \big(\text{top} - \text{bottom}\big)\,dx,$$

with $a$ and $b$ the intersection points (Chapter \@ref(functions)).

No $+C$ is needed — it cancels in the subtraction.

```{r, echo=FALSE}
embed_desmos("y4co25dogp")
```

### Worked examples

**Example (simple).** Evaluate $\displaystyle\int_1^3 x^2\,dx$.

$$\left[\frac{x^3}{3}\right]_1^3 = 9 - \frac{1}{3} = \frac{26}{3}.$$

**Example (moderate — the signed-area trap).** Find (a) $\displaystyle\int_0^3 (x^2 - 4x + 3)\,dx$, and (b) the total **area** enclosed between the curve and the $x$-axis over $[0, 3]$.

(a) $\left[\dfrac{x^3}{3} - 2x^2 + 3x\right]_0^3 = 9 - 18 + 9 = 0.$

(b) Zero area? Of course not. The curve crosses the axis at $x = 1$ and $x = 3$ (factorise: $(x-1)(x-3)$), dipping *below* between them. Piece by piece:

$$\int_0^1 = \frac{4}{3} \quad (\text{above the axis}), \qquad \int_1^3 = -\frac{4}{3} \quad (\text{below}).$$

The signed contributions cancel — that's the zero in (a). The actual area is $\tfrac{4}{3} + \left|-\tfrac{4}{3}\right| = \tfrac{8}{3}$. Moral: for areas, *always sketch and split at the roots*.

**Example (moderate — area between curves).** Find the area enclosed by $y = x^2$ and $y = x + 2$.

Intersections: $x^2 = x + 2 \Rightarrow (x-2)(x+1) = 0$, so $x = -1, 2$. The line is on top between them:

$$\int_{-1}^{2} \big(x + 2 - x^2\big)\,dx = \left[\frac{x^2}{2} + 2x - \frac{x^3}{3}\right]_{-1}^{2} = \frac{10}{3} - \left(-\frac{7}{6}\right) = \frac{9}{2}.$$

**Example (challenging — consumer surplus).** A market has demand curve $p = 100 - 2q$ (the price at which quantity $q$ can be sold) and the market price settles at $p = 40$. **Consumer surplus** is the total amount consumers *would have been willing to pay* above what they actually paid — the area between the demand curve and the price line. Compute it.

At $p = 40$, the quantity sold is $q^* = 30$. Surplus:

$$\text{CS} = \int_0^{30} (100 - 2q)\,dq - 40 \times 30 = \Big[100q - q^2\Big]_0^{30} - 1200 = (3000 - 900) - 1200 = 900.$$

Consumers collectively gain £900 of value beyond their spending. This is a definite integral wearing an economics badge, and it appears — along with its twin, *producer surplus* — in every microeconomics course. Welfare analysis of taxes, price caps and monopolies is done by comparing exactly these areas.

```{r consumer-surplus, echo=FALSE, fig.cap="Consumer surplus: the shaded triangle between the demand curve and the market price.", fig.height=4}
curve(100 - 2*x, from = 0, to = 50, ylab = "Price p", xlab = "Quantity q",
      col = "steelblue", lwd = 2, ylim = c(0, 100))
qq <- seq(0, 30, length.out = 50)
polygon(c(0, qq, 30, 0), c(100, 100 - 2*qq, 40, 40), col = "#B0C4DE80", border = NA)
abline(h = 40, lty = 2, col = "grey40")
segments(30, 0, 30, 40, lty = 3, col = "grey60")
text(10, 65, "CS = £900")
```

### Practice exercises

1. Evaluate $\displaystyle\int_0^2 (3x^2 + 1)\,dx$.
2. Find the area enclosed by $y = x^2$ and the horizontal line $y = 4$.
3. Demand is $p = 80 - q$ and the market price is £50. Find the consumer surplus.
4. Evaluate $\displaystyle\int_1^4 \left(\sqrt{x} - \frac{1}{x}\right) dx$, leaving your answer exact.

**Common mistakes:**

- Integrating across a sign change and calling the result an "area". Sketch, find the roots, split.
- Subtracting the limits the wrong way: it's $F(b) - F(a)$, top limit first.
- In between-curves problems, choosing "top minus bottom" by guesswork — check with a test value or a sketch, or your area will come out negative.

```{r, echo=FALSE}
embed_desmos("z3x0wrhxij")
```

```{r, echo=FALSE}
embed_youtube("M2f-o3OK0W8", 
              title = "The area between two curves")
```


## Integration as accumulation {#accumulation}

### Key idea

The deepest way to see a definite integral: chop $[a, b]$ into slivers of width $\Delta x$, approximate the quantity in each sliver by $f(x)\,\Delta x$, add up, and let the slivers shrink. The integral is the *limit of these sums* — which is why the symbol $\int$ is a stretched S, for sum.

The practical payoff: **if $f(t)$ is a rate, $\displaystyle\int_a^b f(t)\,dt$ is the total accumulated between $t = a$ and $t = b$.** Units multiply: a rate in £/month integrated over months gives £; litres/minute over minutes gives litres.

$$\text{total change in } F \text{ from } a \text{ to } b = \int_a^b F'(t)\,dt.$$

### Worked examples

**Example (simple).** A machine draws power at the rate $r(t) = 5 + 2t$ kW, $t$ hours after switch-on. How much energy does it use in the first 4 hours?

$$\int_0^4 (5 + 2t)\,dt = \Big[5t + t^2\Big]_0^4 = 20 + 16 = 36 \text{ kWh}.$$

**Example (moderate — total sales from a decaying sales rate).** After a rival launches, a product sells at the rate $s(t) = 200\,e^{-0.1t}$ units per month. How many units are sold in the first year — and how does that compare with the naive estimate "200 a month for 12 months"?

$$\int_0^{12} 200\,e^{-0.1t}\,dt = \Big[-2000\,e^{-0.1t}\Big]_0^{12} = 2000\big(1 - e^{-1.2}\big) \approx 1398 \text{ units}.$$

The naive estimate of 2{,}400 ignores the decay and overshoots by 70%. Integrating the *rate* is how you total a changing flow correctly.

**Example (challenging — cumulative profit with losses and gains).** A project's profit *rate* is $P'(t) = -t^2 + 8t - 12$ (£k per month) over its 8-month life. Analyse the project.

Factorise: $P'(t) = -(t-2)(t-6)$: the project **loses** money for $t < 2$ (start-up phase), **makes** money for $2 < t < 6$, then loses again as the market fades. Accumulations:

$$\int_0^2 P'\,dt = -\tfrac{32}{3}, \qquad \int_2^6 P'\,dt = +\tfrac{32}{3}, \qquad \int_6^8 P'\,dt = -\tfrac{32}{3}.$$

The profitable middle exactly repays the start-up losses — and then the final two months throw it away: overall, $\int_0^8 P'\,dt = -\tfrac{32}{3} \approx -£10.7$k. Managerial reading: *stop the project at $t = 6$*, when the profit rate turns negative, and the project breaks even instead of losing £10.7k. The optimal stopping time is where the rate changes sign — a derivative fact — and the consequences are measured by integrals. The two halves of calculus, working one problem together.

### Practice exercises

1. Water flows into a tank at $r(t) = 4 + t$ litres per minute. How much water enters in the first 6 minutes?
2. Sales run at $s(t) = 300\,e^{-0.2t}$ units per month. Find total sales over the first 10 months, to the nearest unit.
3. A firm's spending rate is $c(t)$, measured in £ per month, with $t$ in months. In one sentence, interpret $\displaystyle\int_0^{12} c(t)\,dt$ — including its units.

**Common mistakes:**

- Confusing the rate at a moment with the total over a period: $s(12)$ is the sales *rate* in month 12; $\int_0^{12} s\,dt$ is *total* sales so far. Different objects, different units.
- Multiplying a changing rate by the whole time interval (the "naive 2,400" above). That only works when the rate is constant.
- Dropping units. If the units of your answer don't come out right, the model or the calculation is wrong.

**Why this matters:** Accumulation thinking runs through operations (total throughput from a production rate), finance (accumulated interest and discounted cash flows), and — most importantly for your degree — **probability**: probabilities are areas under density curves, and expected values are integrals. When your statistics lecturer writes $\int x f(x)\,dx$, this section is what it means.

## Integration by substitution {#substitution}

### Key idea

Substitution is the **chain rule in reverse**. If the integrand contains a "lump" $u = g(x)$ *and* (up to a constant) its derivative $g'(x)$ as a factor, substitute:

$$u = g(x), \qquad du = g'(x)\,dx,$$

rewrite everything in terms of $u$, integrate, and substitute back. Two special cases worth automating:

- **Linear insides:** $\displaystyle\int (ax+b)^n dx = \frac{(ax+b)^{n+1}}{a(n+1)} + C$ and $\displaystyle\int e^{ax+b}\,dx = \frac{e^{ax+b}}{a} + C$ — "integrate the outside, divide by the inside's derivative".
- **Definite integrals:** change the limits into $u$-values and never return to $x$ at all.

### Worked examples

**Example (simple).** Find $\displaystyle\int (3x + 2)^4\,dx$.

Linear inside: $\dfrac{(3x+2)^5}{5 \times 3} + C = \dfrac{(3x+2)^5}{15} + C$.

**Example (simple).** Find $\displaystyle\int e^{-0.2t}\,dt$.

$$\frac{e^{-0.2t}}{-0.2} + C = -5\,e^{-0.2t} + C.$$

(The workhorse of every decaying-rate total, as in Section \@ref(accumulation).)

**Example (moderate).** Find $\displaystyle\int 2x\,(x^2 + 1)^5\,dx$.

Spot the pattern: the lump is $x^2 + 1$ and its derivative $2x$ is sitting right there. Let $u = x^2 + 1$, $du = 2x\,dx$:

$$\int u^5\,du = \frac{u^6}{6} + C = \frac{(x^2+1)^6}{6} + C.$$

**Example (moderate).** Find $\displaystyle\int x\,e^{x^2}\,dx$ and $\displaystyle\int \frac{\ln x}{x}\,dx$.

First: $u = x^2$, $du = 2x\,dx$, so $x\,dx = \tfrac12 du$: the integral is $\tfrac{1}{2}e^{x^2} + C$.

Second: $u = \ln x$, $du = \tfrac{1}{x}dx$: the integral is $\displaystyle\int u\,du = \frac{(\ln x)^2}{2} + C$.

In both, the give-away was a factor that is the derivative of the lump ($x$ vs $x^2$; $\tfrac1x$ vs $\ln x$). Train your eye to hunt for that pairing.

**Example (challenging — a definite integral with changed limits).** Evaluate $\displaystyle\int_0^2 x\sqrt{x^2 + 1}\,dx$.

Let $u = x^2 + 1$, $du = 2x\,dx$. **Change the limits:** $x = 0 \Rightarrow u = 1$; $x = 2 \Rightarrow u = 5$:

$$\int_0^2 x\sqrt{x^2+1}\,dx = \frac{1}{2}\int_1^5 u^{1/2}\,du = \frac{1}{2}\left[\frac{2u^{3/2}}{3}\right]_1^5 = \frac{1}{3}\big(5\sqrt{5} - 1\big) \approx 3.39.$$

With the limits converted, there's no need to substitute back — cleaner and less error-prone.

### Practice exercises

1. Find $\displaystyle\int (2x - 7)^5\,dx$.
2. Find $\displaystyle\int 6x\,(x^2 + 4)^2\,dx$.
3. Find $\displaystyle\int x\,e^{-x^2}\,dx$.
4. Evaluate $\displaystyle\int_0^1 x\,(x^2 + 3)^3\,dx$.

**Common mistakes:**

- Forgetting to transform $dx$: $du = g'(x)\,dx$ is part of the substitution, not decoration. If a stray $x$ survives into the $u$-integral, something's gone wrong.
- In definite integrals, substituting $u$ but keeping the old $x$-limits — the single most common substitution error. Change the limits or change back to $x$; never mix.
- Using the linear-inside shortcut on non-linear insides: $\int (x^2+1)^5 dx \ne \tfrac{(x^2+1)^6}{6 \cdot 2x}$ — you cannot "divide by the derivative" when the derivative isn't constant.

```{r, echo=FALSE}
embed_youtube("ahKY8m2B_y", 
              title = "Integration by substitution",
              playlist = "PLg2tfDG3Ww4vgWOxMajRzNuBNju-Liix2")
```

::: {.webex-check .webex-box}
**Self-check: substitution**

**Q1.** Linear inside: $\displaystyle\int (2x+5)^7\,dx = \dfrac{(2x+5)^8}{n} + C$. $n =$ `r fitb(16, num = TRUE)`

**Q2.** Linear inside: $\displaystyle\int e^{-0.4t}\,dt = k\,e^{-0.4t} + C$. $k =$ `r fitb(-2.5, tol = 0.01)`

**Q3.** Which equals $\displaystyle\int x\,(x^2+3)^4\,dx$?

(A) $\dfrac{(x^2+3)^5}{5} + C$  
(B) $\dfrac{(x^2+3)^5}{10} + C$  
(C) $\dfrac{(x^2+3)^5}{10x} + C$  
(D) $\dfrac{x\,(x^2+3)^5}{10} + C$  

Answer: `r mcq(c("A", answer = "B", "C", "D"))`

**Q4.** Which equals $\displaystyle\int \dfrac{(\ln x)^3}{x}\,dx$?

(A) $\dfrac{(\ln x)^4}{4} + C$  
(B) $\dfrac{(\ln x)^4}{4x} + C$  
(C) $\dfrac{1}{4}\ln\big(x^4\big) + C$  
(D) $\dfrac{3(\ln x)^2}{x} + C$  

Answer: `r mcq(c(answer = "A", "B", "C", "D"))`

**Q5.** For $\displaystyle\int_0^1 2x\,(x^2+1)^3\,dx$ with $u = x^2 + 1$: the new $u$-limits are `r mcq(c("0 and 1", answer = "1 and 2", "0 and 2"))`, and the value of the integral is `r fitb(3.75, tol = 0.01)`
:::

`r hide("Show solutions")`
**Q1.** $n = a(n_{\text{new}}) = 2 \times 8 = 16$: integrate the outside, then divide by the inside's derivative. Dividing by 8 alone forgets that inner 2.

**Q2.** $k = \dfrac{1}{-0.4} = -2.5$ — the workhorse of every decaying-rate total.

**Q3.** (B). With $u = x^2+3$, $du = 2x\,dx$, so $x\,dx = \tfrac12\,du$ and the integral is $\tfrac12 \cdot \tfrac{u^5}{5} = \tfrac{u^5}{10}$. (C) is the illegal "divide by the derivative" shortcut — only *constant* derivatives can be divided out; (A) drops the $\tfrac12$; (D) leaves a stray $x$ behind, a sign the substitution wasn't completed. Verify (B): $\tfrac{d}{dx}\tfrac{(x^2+3)^5}{10} = \tfrac{5(x^2+3)^4 \cdot 2x}{10} = x(x^2+3)^4$ ✓.

**Q4.** (A). The give-away pairing: the lump is $\ln x$ and its derivative $\tfrac1x$ is sitting there as a factor. So $\displaystyle\int u^3\,du = \tfrac{u^4}{4}$. (D) is the *derivative* of $(\ln x)^3$; (C) is a log-law muddle — $\tfrac14\ln(x^4) = \ln x$, nothing like $\tfrac{(\ln x)^4}{4}$.

**Q5.** Limits 1 and 2: $u(0) = 1$, $u(1) = 2$ — keeping the old $x$-limits is the single most common substitution error. Then $\displaystyle\int_1^2 u^3\,du = \left[\tfrac{u^4}{4}\right]_1^2 = \tfrac{16-1}{4} = 3.75$, with no need to return to $x$ at all.
`r unhide()`

## Integration by parts {#byparts}

### Key idea

Integration by parts is the **product rule in reverse**:

$$\int u\,\frac{dv}{dx}\,dx = uv - \int v\,\frac{du}{dx}\,dx.$$

Use it for products where substitution fails — classically $x\,e^{kx}$ and $x\ln x$. Choose $u$ to be the factor that gets *simpler* when differentiated (usually the polynomial; for $\ln x$ integrals, choose $u = \ln x$ since it differentiates to the tame $\tfrac1x$). The remaining factor is $\tfrac{dv}{dx}$, which you must be able to integrate.

### Worked examples

**Example (simple).** Find $\displaystyle\int x\,e^x\,dx$.

Take $u = x$ (differentiates to 1 — simpler) and $\tfrac{dv}{dx} = e^x$ (so $v = e^x$):

$$\int x e^x\,dx = x e^x - \int e^x\,dx = x e^x - e^x + C = e^x(x - 1) + C.$$

**Example (moderate).** Find $\displaystyle\int x\,e^{-2x}\,dx$.

$u = x$, $\tfrac{dv}{dx} = e^{-2x}$, so $v = -\tfrac{1}{2}e^{-2x}$:

$$\int x e^{-2x}\,dx = -\frac{x}{2}e^{-2x} + \frac{1}{2}\int e^{-2x}\,dx = -\frac{x}{2}e^{-2x} - \frac{1}{4}e^{-2x} + C = -\frac{e^{-2x}(2x+1)}{4} + C.$$

(Watch the signs — this integral is a sign-error factory. Differentiate your answer to check.)

**Example (moderate — the famous trick).** Find $\displaystyle\int \ln x\,dx$.

It doesn't look like a product — *make* it one: $u = \ln x$, $\tfrac{dv}{dx} = 1$ (so $v = x$):

$$\int \ln x\,dx = x\ln x - \int x \cdot \frac{1}{x}\,dx = x\ln x - x + C.$$

**Example (challenging — totalling a sales pulse).** A promotional campaign produces sales at the rate $s(t) = 50\,t\,e^{-0.5t}$ units per week (a rapid rise, then a long fade — the shape you differentiated in Section \@ref(productrule)). Find the total sales over the first 4 weeks.

We need $\displaystyle 50\int_0^4 t\,e^{-0.5t}\,dt$. By parts with $u = t$, $v = -2e^{-0.5t}$:

$$\int t\,e^{-0.5t}\,dt = -2t\,e^{-0.5t} + 2\int e^{-0.5t}\,dt = -2t\,e^{-0.5t} - 4e^{-0.5t}.$$

Evaluate from 0 to 4:

$$\Big[-2t e^{-0.5t} - 4e^{-0.5t}\Big]_0^4 = \big(-8e^{-2} - 4e^{-2}\big) - (-4) = 4 - 12e^{-2} \approx 2.376.$$

Total sales $\approx 50 \times 2.376 \approx 119$ units. (Rise-and-fade response curves like $t\,e^{-kt}$ are standard in advertising models, and by parts is how you total them.)

### Practice exercises

1. Find $\displaystyle\int x\,e^{3x}\,dx$.
2. Find $\displaystyle\int x\ln x\,dx$.
3. Evaluate $\displaystyle\int_0^1 x\,e^{-x}\,dx$, exactly and to 3 d.p.

**Common mistakes:**

- Choosing $u$ and $dv$ the wrong way round, which makes the new integral *harder*: if $\int v\,\tfrac{du}{dx}\,dx$ looks worse than what you started with, swap and restart.
- Sign errors with negative exponentials — the minus signs multiply up. Check by differentiating.
- Forgetting that $v$ must come from *integrating* $\tfrac{dv}{dx}$: with $\tfrac{dv}{dx} = e^{-2x}$, $v = -\tfrac12 e^{-2x}$, not $e^{-2x}$.

```{r, echo=FALSE}
embed_youtube("MeHuy4OAuus", 
              title = "Integration by parts",
              playlist = "PLg2tfDG3Ww4vgWOxMajRzNuBNju-Liix2&index=26")
```

::: {.webex-check .webex-box}
**Self-check: integration by parts (and one impostor)**

**Q1.** Which equals $\displaystyle\int x\,e^{2x}\,dx$?

(A) $\dfrac{e^{2x}(2x+1)}{4} + C$  
(B) $x\,e^{2x} - \tfrac{1}{2}e^{2x} + C$  
(C) $\dfrac{e^{2x}(2x-1)}{4} + C$  
(D) $\dfrac{x^2 e^{2x}}{4} + C$  

Answer: `r mcq(c("A", "B", answer = "C", "D"))`

**Q2.** One integral in this quiz is an impostor. For $\displaystyle\int x\,e^{x^2}\,dx$ the right technique is `r mcq(c("integration by parts", answer = "substitution", "partial fractions"))`, and $\displaystyle\int x\,e^{x^2}\,dx = k\,e^{x^2} + C$ with $k =$ `r fitb(0.5, tol = 0.01)`

**Q3.** Which equals $\displaystyle\int x\ln x\,dx$?

(A) $\dfrac{x^2\ln x}{2} + \dfrac{x^2}{4} + C$  
(B) $\dfrac{x^2\ln x}{2} - \dfrac{x^2}{2} + C$  
(C) $x\ln x - x + C$  
(D) $\dfrac{x^2\ln x}{2} - \dfrac{x^2}{4} + C$  

Answer: `r mcq(c("A", "B", "C", answer = "D"))`

**Q4.** The famous trick: $\displaystyle\int \ln x\,dx = x\ln x + a\,x + C$. $a =$ `r fitb(-1, num = TRUE)`

**Q5.** A campaign generates sales at the rate $s(t) = 40\,t\,e^{-t}$ units per week. Total sales over the first 2 weeks, to 1 d.p.: `r fitb(23.8, tol = 0.1)` units.
:::

`r hide("Show solutions")`
**Q1.** (C). Take $u = x$, $\tfrac{dv}{dx} = e^{2x}$ so $v = \tfrac12 e^{2x}$: $\tfrac{x e^{2x}}{2} - \tfrac12\displaystyle\int e^{2x}\,dx = \tfrac{xe^{2x}}{2} - \tfrac{e^{2x}}{4} = \tfrac{e^{2x}(2x-1)}{4}$. (B) used $v = e^{2x}$ — but $v$ must come from *integrating* $\tfrac{dv}{dx}$; (A) is the sign slip; (D) integrates the product factor by factor, which is never legal. Verify (C): differentiating gives $\tfrac{2e^{2x}(2x-1) + 2e^{2x}}{4} = xe^{2x}$ ✓.

**Q2.** Substitution — the factor $x$ is (up to the constant 2) the derivative of the lump $x^2$, the tell-tale pairing. Then $k = \tfrac12$. Trying parts here just sends the integral round in circles; always scan for the lump-and-derivative pattern *before* reaching for parts.

**Q3.** (D). Take $u = \ln x$ (it differentiates to the tame $\tfrac1x$), $\tfrac{dv}{dx} = x$: $\tfrac{x^2\ln x}{2} - \displaystyle\int \tfrac{x}{2}\,dx = \tfrac{x^2\ln x}{2} - \tfrac{x^2}{4}$. (B) forgets the second division by 2 in $\int\tfrac{x}{2}\,dx$; (C) is the answer to a *different* integral, $\int \ln x\,dx$.

**Q4.** $a = -1$: with $u = \ln x$ and $\tfrac{dv}{dx} = 1$, $\displaystyle\int \ln x\,dx = x\ln x - \int x \cdot \tfrac1x\,dx = x\ln x - x + C$.

**Q5.** $\displaystyle\int t\,e^{-t}\,dt = -(t+1)e^{-t}$, so $\displaystyle\int_0^2 t\,e^{-t}\,dt = 1 - 3e^{-2} \approx 0.594$ and total sales $\approx 40 \times 0.594 \approx 23.8$ units. Watch the multiplying minus signs — this shape is a sign-error factory, so differentiate your antiderivative to check.
`r unhide()`

## Integration using partial fractions {#pf-integration}

### Key idea

Here is the payoff promised in Section \@ref(partialfractions): a rational function with linear factors splits into simple fractions, and each piece integrates to a logarithm:

$$\int \frac{A}{x + a}\,dx = A\ln|x + a| + C, \qquad \int \frac{B}{(x+a)^2}\,dx = -\frac{B}{x+a} + C.$$

The routine: **decompose (Chapter \@ref(algebra) algebra), then integrate term by term.**

### Worked examples

**Example (simple).** Find $\displaystyle\int \frac{3x + 5}{(x+1)(x+3)}\,dx$.

Section \@ref(partialfractions) already did the decomposition: $\dfrac{1}{x+1} + \dfrac{2}{x+3}$. So

$$\int \frac{3x+5}{(x+1)(x+3)}\,dx = \ln|x+1| + 2\ln|x+3| + C.$$

**Example (moderate).** Find $\displaystyle\int \frac{1}{x(x - 2)}\,dx$.

Decompose: $\dfrac{1}{x(x-2)} = \dfrac{A}{x} + \dfrac{B}{x-2}$ gives $A = -\tfrac12$, $B = \tfrac12$:

$$\int \frac{dx}{x(x-2)} = \frac{1}{2}\ln|x - 2| - \frac{1}{2}\ln|x| + C = \frac{1}{2}\ln\left|\frac{x-2}{x}\right| + C.$$

(Compressing with the log laws of Section \@ref(loglaws) is optional but tidy.)

**Example (moderate — a repeated factor).** Find $\displaystyle\int \frac{3x + 1}{(x-1)^2}\,dx$.

From Section \@ref(partialfractions): $\dfrac{3}{x-1} + \dfrac{4}{(x-1)^2}$. Integrate — note the second piece is a *power*, not a log:

$$3\ln|x - 1| - \frac{4}{x-1} + C.$$

**Example (challenging — a definite integral, and a look ahead).** (a) Evaluate $\displaystyle\int_3^4 \frac{3x+5}{(x+1)(x+3)}\,dx$ exactly. (b) Why do integrals of the shape $\displaystyle\int \frac{dS}{S(M - S)}$ matter?

(a) Using the antiderivative above:

$$\Big[\ln(x+1) + 2\ln(x+3)\Big]_3^4 = \big(\ln 5 + 2\ln 7\big) - \big(\ln 4 + 2\ln 6\big) = \ln\frac{5}{4} + 2\ln\frac{7}{6} \approx 0.531.$$

(No absolute values needed — every argument is positive on $[3,4]$.)

(b) That shape appears when solving the **logistic differential equation** $\tfrac{dS}{dt} = kS(M - S)$, the standard model of product adoption and epidemic-style growth ("slow start, rapid middle, saturating finish"). Separating it — next section's technique — leaves exactly $\int \tfrac{dS}{S(M-S)}$, which partial fractions cracks open: $\tfrac{1}{S(M-S)} = \tfrac{1}{M}\big(\tfrac{1}{S} + \tfrac{1}{M-S}\big)$. Your A-level algebra is the key to a genuinely important business model.

### Practice exercises

1. Find $\displaystyle\int \frac{7x + 18}{(x+2)(x+4)}\,dx$ (the decomposition was Exercise 1 of Section \@ref(partialfractions)).
2. Find $\displaystyle\int \frac{5}{(x-1)(x+4)}\,dx$.
3. Evaluate $\displaystyle\int_0^1 \frac{1}{(x+1)(x+2)}\,dx$ exactly, and simplify using the log laws.

**Common mistakes:**

- Skipping the decomposition and inventing $\int \tfrac{f}{g} = \ln|g|$-style "rules". Only $\int \tfrac{g'(x)}{g(x)}dx = \ln|g(x)| + C$ is legitimate — check the numerator really is the denominator's derivative.
- Integrating $\tfrac{B}{(x+a)^2}$ to a logarithm. Repeated-factor pieces are negative powers: they integrate by the power rule.
- Dropping the modulus signs in $\ln|x+a|$ for indefinite integrals.

::: {.webex-check .webex-box}
**Self-check: partial fractions and integration**

**Q1.** Decompose $\dfrac{5x+7}{(x+1)(x+2)} = \dfrac{A}{x+1} + \dfrac{B}{x+2}$. $A =$ `r fitb(2, num = TRUE)` and $B =$ `r fitb(3, num = TRUE)`

**Q2.** Hence, which equals $\displaystyle\int \dfrac{5x+7}{(x+1)(x+2)}\,dx$?

(A) $\ln\big|x^2 + 3x + 2\big| + C$  
(B) $2\ln|x+1| + 3\ln|x+2| + C$  
(C) $3\ln|x+1| + 2\ln|x+2| + C$  
(D) $\dfrac{2}{x+1} + \dfrac{3}{x+2} + C$  

Answer: `r mcq(c("A", answer = "B", "C", "D"))`

**Q3.** For $\displaystyle\int \dfrac{2x+3}{x^2+3x+2}\,dx$, the quickest route is to `r mcq(c("decompose into partial fractions", answer = "spot that the numerator is the derivative of the denominator", "integrate numerator and denominator separately"))`, giving $k\ln\big|x^2+3x+2\big| + C$ with $k =$ `r fitb(1, num = TRUE)`

**Q4.** Which equals $\displaystyle\int \dfrac{2x+5}{(x+2)^2}\,dx$?

(A) $2\ln|x+2| + \dfrac{1}{x+2} + C$  
(B) $2\ln|x+2| - \dfrac{1}{(x+2)^2} + C$  
(C) $2\ln|x+2| + \ln\big((x+2)^2\big) + C$  
(D) $2\ln|x+2| - \dfrac{1}{x+2} + C$  

Answer: `r mcq(c("A", "B", "C", answer = "D"))`

**Q5.** $\displaystyle\int_0^2 \dfrac{1}{(x+1)(x+3)}\,dx = \tfrac{1}{2}\ln k$ exactly. $k$ (as a decimal) $=$ `r fitb(1.8, tol = 0.01)`
:::

`r hide("Show solutions")`
**Q1.** Cover-up: at $x = -1$, $A = \tfrac{-5+7}{1} = 2$; at $x = -2$, $B = \tfrac{-10+7}{-1} = 3$.

**Q2.** (B): each simple fraction becomes a log with its coefficient. (A) is the invented rule "$\int \tfrac{f}{g} = \ln|g|$" — legitimate *only* when the numerator is exactly the denominator's derivative, which here would need $2x+3$, not $5x+7$. (C) swaps $A$ and $B$; (D) forgot to integrate at all.

**Q3.** Here the numerator $2x + 3$ *is* $\tfrac{d}{dx}(x^2+3x+2)$, so $\int \tfrac{g'}{g}\,dx = \ln|g| + C$ applies directly and $k = 1$ — no decomposition needed. Always check for this shortcut before decomposing.

**Q4.** (D). Decompose: $\dfrac{2x+5}{(x+2)^2} = \dfrac{2}{x+2} + \dfrac{1}{(x+2)^2}$, and the repeated-factor piece is a *power*, not a log: $\displaystyle\int (x+2)^{-2}\,dx = -(x+2)^{-1}$. (C) logs the repeated piece — the trap named in the Common mistakes box above; (A) drops the minus sign from the power rule; (B) forgets to raise the power.

**Q5.** $\dfrac{1}{(x+1)(x+3)} = \tfrac12\left(\dfrac{1}{x+1} - \dfrac{1}{x+3}\right)$, so the integral is $\tfrac12\Big[\ln\tfrac{x+1}{x+3}\Big]_0^2 = \tfrac12\big(\ln\tfrac35 - \ln\tfrac13\big) = \tfrac12\ln\tfrac95$, giving $k = 1.8$. (All arguments are positive on $[0,2]$, so no modulus signs are needed here.)
`r unhide()`

## Separable differential equations {#separable-des}

### Key idea

Section \@ref(constructing-des) taught you to *write* differential equations; now we solve the separable ones. A first-order DE is **separable** if it can be rearranged so each variable sits with its own differential:

$$\frac{dy}{dx} = f(x)\,g(y) \;\;\Longrightarrow\;\; \int \frac{dy}{g(y)} = \int f(x)\,dx.$$

Integrate both sides, keep **one** constant $C$, then use the initial condition to find it (giving the **particular solution**). Expect to finish with algebra: exponentiating to undo logs, tidying constants ($e^{C}$ becomes a new constant $A$).

### Worked examples

**Example (simple — exponential growth, derived at last).** Solve $\dfrac{dP}{dt} = 0.03P$ with $P(0) = 50$.

Separate and integrate:

$$\int \frac{dP}{P} = \int 0.03\,dt \;\Rightarrow\; \ln P = 0.03t + C.$$

Exponentiate: $P = e^{0.03t + C} = A\,e^{0.03t}$ where $A = e^C$. The condition $P(0) = 50$ gives $A = 50$:

$$P(t) = 50\,e^{0.03t}.$$

The exponential growth of Chapter \@ref(explog) is no longer an assumption — it is the *unique consequence* of "grows at a rate proportional to its size".

**Example (moderate — exponential depreciation, derived).** Solve $\dfrac{dV}{dt} = -0.15V$ with $V(0) = 20000$, and find when the value halves.

Same separation: $V(t) = 20000\,e^{-0.15t}$. Halving: $e^{-0.15t} = \tfrac12$, so $t = \dfrac{\ln 2}{0.15} \approx 4.6$ years — the machine from the Chapter \@ref(chainrule) exercises, now with its model built from first principles.

**Example (challenging — saturating growth, start to finish).** A streaming service estimates its saturation level at $M = 5{,}000$ (thousand subscribers), and models growth by $\dfrac{dS}{dt} = k(M - S)$ with $S(0) = 0$ (Section \@ref(constructing-des)). (a) Solve the DE. (b) Given $S(2) = 1500$, find $k$. (c) When does the service reach 4{,}000?

(a) Separate:

$$\int \frac{dS}{M - S} = \int k\,dt \;\Rightarrow\; -\ln|M - S| = kt + C.$$

So $M - S = A\,e^{-kt}$; the condition $S(0) = 0$ gives $A = M$, hence

$$S(t) = M\big(1 - e^{-kt}\big).$$

(b) $1500 = 5000\big(1 - e^{-2k}\big) \Rightarrow e^{-2k} = 0.7 \Rightarrow k = -\tfrac{\ln 0.7}{2} \approx 0.178$ per year.

(c) $4000 = 5000\big(1 - e^{-kt}\big) \Rightarrow e^{-kt} = 0.2 \Rightarrow t = \dfrac{\ln 5}{0.178} \approx 9.0$ years.

Look at the shape of the solution: fast early growth, then an ever-slower crawl towards (but never reaching) the ceiling $M$ — realistic saturation, produced by three lines of separation and one initial condition. The first 30% of the market took 2 years; the *next* 50% takes 7 more.

```{r saturating-growth, echo=FALSE, fig.cap="The solution S(t) = M(1 - e^{-kt}): rapid early growth saturating towards the ceiling M.", fig.height=4}
k <- -log(0.7) / 2
curve(5000 * (1 - exp(-k * x)), from = 0, to = 20, ylab = "Subscribers (000s)",
      xlab = "Years", col = "steelblue", lwd = 2, ylim = c(0, 5500))
abline(h = 5000, lty = 2, col = "firebrick")
points(c(2, 9.03), c(1500, 4000), pch = 16)
text(16, 5250, "M = 5000", col = "firebrick")
```

### Practice exercises

1. Solve $\dfrac{dy}{dx} = y$ (for $y > 0$) with $y(0) = 3$.
2. Solve $\dfrac{dN}{dt} = 0.4N$ with $N(0) = 200$, and evaluate $N(5)$ to the nearest whole number.
3. Coffee cools according to $\dfrac{dT}{dt} = -0.1(T - 20)$ with $T(0) = 90$ (°C, minutes). Solve for $T(t)$, and find when the coffee reaches 40°C.
4. *(Challenge.)* A pension pot obeys $\dfrac{dB}{dt} = 0.05B - 2000$ with $B(0) = 30000$. Solve the DE (hint: it separates as $\int \tfrac{dB}{0.05B - 2000} = \int dt$; expect a solution of the form $B = 40000 + Ae^{0.05t}$) and find when the pot runs out.

**Common mistakes:**

- Splitting $\tfrac{dy}{dx}$ carelessly: separation must gather *all* $y$'s (including any $g(y)$ factor) on the $dy$ side. If $y$'s remain on the $x$ side, the equation wasn't separated.
- Two constants of integration. Integrate both sides, write **one** $C$ (the difference of the two).
- Applying the initial condition too late — or not at all. The question usually wants the *particular* solution, and every un-found constant is an unanswered question.
- Sloppy exponentiation: from $\ln P = kt + C$, we get $P = e^C e^{kt} = Ae^{kt}$; the constant multiplies, it doesn't add.


```{r, echo=FALSE}
embed_desmos("bhb4zmdp49")
```

## Interpreting integrals and solutions in context {#interpretation}

### Key idea

The final syllabus skill is the one universities value most: saying what your mathematics *means*, and where it stops being trustworthy. Three habits:

1. **Units first.** The units of $\int_a^b f(t)\,dt$ are (units of $f$) × (units of $t$). If they aren't the units the question expects, stop and find the error.
2. **Describe the behaviour of solutions.** Initial value, direction of change, long-run behaviour (limits and asymptotes), and any thresholds or steady states.
3. **State limitations honestly.** Every model in this chapter breaks somewhere: say where, and if possible suggest the refinement.

### Worked examples

**Example (simple).** $r(t)$ is a delivery firm's fuel-use rate in litres per hour, with $t$ in hours since midnight. Interpret $\displaystyle\int_6^{18} r(t)\,dt = 340$.

Between 06:00 and 18:00 the fleet used 340 litres of fuel. (Litres/hour × hours = litres ✓.)

**Example (moderate).** Interpret the solution $S(t) = 5000\big(1 - e^{-0.178t}\big)$ from Section \@ref(separable-des): initial behaviour, long-run behaviour, and the meaning of the constant 5{,}000.

At $t = 0$, $S = 0$ and growth is at its fastest (initial rate $kM \approx 890$ per year); growth then slows continuously; as $t \to \infty$, $S \to 5000$ but never reaches it — the horizontal asymptote (Section \@ref(asymptotes)) *is* the saturation level. The constant 5{,}000 is the model's estimate of the total addressable market.

**Example (challenging — an honest model critique).** The growth model $P(t) = 50\,e^{0.03t}$ fits a firm's customer numbers well for its first three years. Give two distinct reasons to distrust its 20-year forecast, and one concrete refinement.

First, *unbounded growth*: $e^{0.03t}$ grows forever, but customers are drawn from a finite population — every exponential forecast eventually predicts more customers than there are people. Second, *frozen conditions*: the constant $k = 0.03$ hard-codes today's competitive environment, pricing and product; over 20 years, rivals enter and markets shift, so the parameter itself will drift. A concrete refinement: replace $\tfrac{dP}{dt} = kP$ with the saturating $\tfrac{dP}{dt} = k(M - P)$, or the logistic $\tfrac{dP}{dt} = kP(M - P)$ (Section \@ref(pf-integration)), which respect a market ceiling $M$. Model criticism of this kind is not a decoration in Management Science coursework — it's where the marks (and, later, the credibility) live.

### Practice exercises

1. $c(t)$ is a household's electricity use rate in kWh per day, $t$ in days. State the meaning and units of $\displaystyle\int_0^{90} c(t)\,dt$.
2. For $V(t) = 20000\,e^{-0.15t}$: describe the long-run behaviour, and give one reason the model misprices an old machine.
3. Give one reason the saturating model $S = M(1 - e^{-kt})$ might *overestimate* early adoption of a genuinely new product, and name the alternative model shape that fixes it.

**Common mistakes:**

- Reporting a number without units or a sentence. "$\approx 1398$" is not an answer; "about 1{,}400 units sold in the first year" is.
- Treating asymptotes as attained values: the model *approaches* $M$; it doesn't arrive.
- Criticising a model with vague words ("it's not accurate"). Good limitations are specific: what assumption fails, when, and with what consequence.

**Why this matters:** University assessment in Management Science persistently pairs "calculate" with "interpret" and "evaluate the model". The calculation earns half the marks; the sentence about meaning and the sentence about limitations earn the rest. This section is that skill, practised on integrals and DEs — and it transfers to every model you will ever build.

## Chapter summary

You should now be able to:

- reverse differentiation, using initial conditions to pin down the constant of integration (and interpret it — e.g. as a fixed cost);
- integrate powers ($n \ne -1$), exponentials and $\tfrac1x$, rewriting expressions first where needed;
- evaluate definite integrals, handle signed area correctly, find areas between curves, and compute consumer surplus;
- treat integrals as accumulations, totalling changing rates with correct units;
- integrate by substitution (including changed limits) and by parts (including the $\ln x$ trick);
- integrate rational functions via partial fractions, including repeated linear factors;
- solve separable differential equations with initial conditions, deriving exponential and saturating growth;
- interpret integral quantities and DE solutions in context, and criticise models constructively.

Not every integral yields to these techniques — and not every one needs to. Chapter \@ref(numerical) shows how to *approximate* integrals (the trapezium rule) and solve equations numerically when exact methods run out.

Treat this diagnostic as a compass, not a verdict: it shows where to point your revision before Chapter \@ref(numerical), nothing more.

## End of chapter quiz

::: {.webex-check .webex-box}

**Q1.** Technique selection — for each integral, choose the most direct method.

(i) $\displaystyle\int x\,e^{x}\,dx$: `r mcq(c(answer = "by parts", "substitution", "partial fractions", "rewrite, then standard integrals"))`  
(ii) $\displaystyle\int 2x\,(x^2+5)^9\,dx$: `r mcq(c("by parts", answer = "substitution", "partial fractions", "rewrite, then standard integrals"))`  
(iii) $\displaystyle\int \dfrac{x+3}{(x+1)(x+2)}\,dx$: `r mcq(c("by parts", "substitution", answer = "partial fractions", "rewrite, then standard integrals"))`  
(iv) $\displaystyle\int \dfrac{x^3 + 2x}{x}\,dx$: `r mcq(c("by parts", "substitution", "partial fractions", answer = "rewrite, then standard integrals"))`

**Q2.** Given $\dfrac{dy}{dx} = 6x^2$ and $y(1) = 5$, find the constant of integration. $C =$ `r fitb(3, num = TRUE)`

**Q3.** Which equals $\displaystyle\int \dfrac{6}{\sqrt{x}}\,dx$?

(A) $12\sqrt{x} + C$  
(B) $3\sqrt{x} + C$  
(C) $6\ln\sqrt{x} + C$  
(D) $-3x^{-3/2} + C$  

Answer: `r mcq(c(answer = "A", "B", "C", "D"))`

**Q4.** For $y = x^2 - 4x$ on $[0, 4]$: $\displaystyle\int_0^4 (x^2 - 4x)\,dx$, to 2 d.p., is `r fitb(-10.67, tol = 0.02)`, while the total **area** enclosed between the curve and the $x$-axis, to 2 d.p., is `r fitb(10.67, tol = 0.02)`

**Q5.** A market has demand curve $p = 60 - 3q$ and the market price settles at £24. The consumer surplus, in £, is `r fitb(216, num = TRUE)`

**Q6.** A machine draws power at the rate $r(t) = 3 + t$ kW, $t$ hours after switch-on. Over the first 8 hours it uses `r fitb(56, num = TRUE)` units of energy, measured in `r mcq(c("kW", answer = "kWh", "kW per hour"))`.

**Q7.** After a rival launches, sales run at $s(t) = 100\,e^{-0.25t}$ units per month. Total sales over the first 6 months, to the nearest unit: `r fitb(311, tol = 1)`

**Q8.** With $s(t)$ a sales *rate* in units per month, the quantity $\displaystyle\int_0^{12} s(t)\,dt$ is `r mcq(c("the sales rate in month 12", answer = "the total units sold in the first year", "the average monthly sales rate"))`.

**Q9.** Evaluate $\displaystyle\int_0^1 x\,(x^2+1)^4\,dx$ (change the limits!). Value: `r fitb(3.1, tol = 0.01)`

**Q10.** Which equals $\displaystyle\int x\,e^{-x}\,dx$?

(A) $(x+1)e^{-x} + C$  
(B) $-(x+1)e^{-x} + C$  
(C) $(x-1)e^{-x} + C$  
(D) $-\dfrac{x^2 e^{-x}}{2} + C$  

Answer: `r mcq(c("A", answer = "B", "C", "D"))`

**Q11.** Evaluate $\displaystyle\int_1^e \ln x\,dx$ exactly: `r fitb(1, num = TRUE)`

**Q12.** Decompose $\dfrac{x+4}{x(x+2)} = \dfrac{A}{x} + \dfrac{B}{x+2}$. $A =$ `r fitb(2, num = TRUE)` and $B =$ `r fitb(-1, num = TRUE)`

**Q13.** Which equals $\displaystyle\int \dfrac{2x}{x^2+9}\,dx$?

(A) $\dfrac{\ln\big(x^2+9\big)}{2x} + C$  
(B) $\ln|2x| + C$  
(C) $\ln\big(x^2+9\big) + C$  
(D) $x^2\ln\big(x^2+9\big) + C$  

Answer: `r mcq(c("A", "B", answer = "C", "D"))`

**Q14.** *A full modelling problem.* A subscription service estimates its saturation level at $M = 8{,}000$ customers and models growth by $\dfrac{dS}{dt} = k(M - S)$ with $S(0) = 0$ ($t$ in years).

(a) Solving the differential equation with this initial condition gives:

(A) $S = M\,e^{-kt}$  
(B) $S = M\,e^{kt}$  
(C) $S = M\big(1 + e^{-kt}\big)$  
(D) $S = M\big(1 - e^{-kt}\big)$  

Answer: `r mcq(c("A", "B", "C", answer = "D"))`

(b) Given $S(3) = 2{,}000$, find $k$ to 3 d.p.: `r fitb(0.096, tol = 0.002)`

(c) Hence the service reaches $6{,}000$ customers after `r fitb(14.5, tol = 0.2)` years (1 d.p.).

**Q15.** Customer numbers obey $\dfrac{dP}{dt} = 0.05P$ with $P(0) = 400$. To the nearest whole number, $P(10) =$ `r fitb(659, tol = 1)`
:::

`r hide("Show solutions")`
**Q1.** (i) by parts — a product with no lump-and-derivative pairing; (ii) substitution — the factor $2x$ is exactly the derivative of the lump $x^2 + 5$; (iii) partial fractions — a rational function with distinct linear factors; (iv) rewrite — divide through to get $x^2 + 2$, then integrate term by term. Choosing the technique *is* the skill; the mechanics follow. *Shaky? Revisit Sections \@ref(substitution)–\@ref(pf-integration).*

**Q2.** $y = 2x^3 + C$; $y(1) = 2 + C = 5$, so $C = 3$. *Shaky? Revisit Section \@ref(antiderivatives).*

**Q3.** (A). Rewrite as $6x^{-1/2}$, integrate to $6 \cdot 2x^{1/2} = 12\sqrt{x}$. (D) is the derivative; (C) misreads a reciprocal power as the log case — only $x^{-1}$ integrates to a log; (B) divides by 2 instead of multiplying. *Shaky? Revisit Section \@ref(standard-integrals).*

**Q4.** $\left[\tfrac{x^3}{3} - 2x^2\right]_0^4 = \tfrac{64}{3} - 32 = -\tfrac{32}{3} \approx -10.67$. The curve lies *below* the axis on the whole of $(0, 4)$ (roots at 0 and 4), so the enclosed area is $\big|{-\tfrac{32}{3}}\big| \approx 10.67$. A negative "area" is always a signed-integral warning: sketch and split at the roots. *Shaky? Revisit Section \@ref(definite-integrals).*

**Q5.** At $p = 24$, $q^* = 12$. $\text{CS} = \displaystyle\int_0^{12}(60 - 3q)\,dq - 24 \times 12 = \big[60q - 1.5q^2\big]_0^{12} - 288 = 504 - 288 = £216$. *Shaky? Revisit Section \@ref(definite-integrals).*

**Q6.** $\displaystyle\int_0^8 (3 + t)\,dt = \big[3t + \tfrac{t^2}{2}\big]_0^8 = 24 + 32 = 56$ kWh: kW × hours = kWh. If your units don't multiply out correctly, stop and find the error. *Shaky? Revisit Section \@ref(accumulation).*

**Q7.** $\displaystyle\int_0^6 100\,e^{-0.25t}\,dt = \big[-400\,e^{-0.25t}\big]_0^6 = 400\big(1 - e^{-1.5}\big) \approx 311$ units. The naive "100 a month for 6 months = 600" overshoots badly: changing rates must be *integrated*, not multiplied by the interval. *Shaky? Revisit Section \@ref(accumulation).*

**Q8.** Total units sold in the first year: integrating a rate accumulates a total (units/month × months = units). The first option confuses the rate *at* a moment with the total *over* a period — different objects, different units. *Shaky? Revisit Sections \@ref(accumulation) and \@ref(interpretation).*

**Q9.** $u = x^2 + 1$ runs from 1 to 2: $\tfrac12\displaystyle\int_1^2 u^4\,du = \tfrac12\left[\tfrac{u^5}{5}\right]_1^2 = \tfrac{31}{10} = 3.1$. Keeping the old $x$-limits after substituting is the classic error — change the limits or change back, never mix. *Shaky? Revisit Section \@ref(substitution).*

**Q10.** (B). $u = x$, $v = -e^{-x}$: $-xe^{-x} + \displaystyle\int e^{-x}\,dx = -xe^{-x} - e^{-x} = -(x+1)e^{-x}$. (A) and (C) are sign slips — the minus signs multiply up with negative exponentials; (D) integrates factor by factor. Verify (B) by differentiating: $-e^{-x} + (x+1)e^{-x} = xe^{-x}$ ✓. *Shaky? Revisit Section \@ref(byparts).*

**Q11.** $\big[x\ln x - x\big]_1^e = (e - e) - (0 - 1) = 1$. *Shaky? Revisit Section \@ref(byparts).*

**Q12.** Cover-up: at $x = 0$, $A = \tfrac{4}{2} = 2$; at $x = -2$, $B = \tfrac{2}{-2} = -1$. *Shaky? Revisit Section \@ref(pf-integration).*

**Q13.** (C). The numerator $2x$ is exactly the derivative of the denominator, so $\int \tfrac{g'}{g}\,dx = \ln|g| + C$ applies directly — no decomposition, and (A)'s "divide by the derivative" is never legal for non-constant derivatives. *Shaky? Revisit Section \@ref(pf-integration).*

**Q14.** (a) (D). Separating gives $-\ln|M - S| = kt + C$, so $M - S = A\,e^{-kt}$; $S(0) = 0$ forces $A = M$, hence $S = M\big(1 - e^{-kt}\big)$. Only (D) satisfies both $S(0) = 0$ and $S \to M$ — checking the initial condition against the options is a ten-second filter. (b) $2000 = 8000\big(1 - e^{-3k}\big) \Rightarrow e^{-3k} = 0.75 \Rightarrow k = -\tfrac{\ln 0.75}{3} \approx 0.096$ per year. (c) $6000 = 8000\big(1 - e^{-kt}\big) \Rightarrow e^{-kt} = 0.25 \Rightarrow t = \tfrac{\ln 4}{k} \approx 14.5$ years. Note the saturation shape: the first quarter of the market took 3 years; reaching three-quarters takes almost fifteen. *Shaky? Revisit Section \@ref(separable-des).*

**Q15.** $P = 400\,e^{0.05t}$, so $P(10) = 400\,e^{0.5} \approx 659$. Remember the exponentiation step: from $\ln P = 0.05t + C$, the constant becomes a *multiplier*, $P = Ae^{0.05t}$. *Shaky? Revisit Section \@ref(separable-des).*
`r unhide()`

Rule of thumb: if two or three questions from the same section went wrong, rework that section before starting Chapter \@ref(numerical). One slip is noise; a cluster is a signal.

## Answers to practice exercises {#integration-answers}

**Antiderivatives:** 1. $2x^4 - x^3 + 2x + C$.  2. $y = 6\sqrt{x} + C$; $y(9) = 18 + C = 20$, so $y = 6\sqrt{x} + 2$.  3. $R(q) = 50q - 0.2q^2$.

**Standard integrals:** 1. $x^4 - 2x^3 + 2x + C$.  2. $12\sqrt{x} + \dfrac{2}{x^2} + C$.  3. $\displaystyle\int \left(x + \frac{4}{x}\right)dx = \frac{x^2}{2} + 4\ln|x| + C$.  4. $5e^x - 3\ln|x| + C$.

**Definite integrals and area:** 1. $[x^3 + x]_0^2 = 10$.  2. Intersections at $x = \pm 2$: $\displaystyle\int_{-2}^{2}(4 - x^2)\,dx = \frac{32}{3}$.  3. $q^* = 30$; $\displaystyle\int_0^{30}(80 - q)\,dq = \Big[80q - \tfrac{q^2}{2}\Big]_0^{30} = 2400 - 450 = 1950$, so $\text{CS} = 1950 - 50 \times 30 = £450$.  4. $\left[\tfrac{2x^{3/2}}{3} - \ln x\right]_1^4 = \tfrac{16}{3} - \ln 4 - \tfrac{2}{3} = \tfrac{14}{3} - \ln 4$.

**Accumulation:** 1. $[4t + \tfrac{t^2}{2}]_0^6 = 24 + 18 = 42$ litres.  2. $1500(1 - e^{-2}) \approx 1297$ units.  3. Total spending over the first year, in pounds (£/month × months = £).

**Substitution:** 1. $\dfrac{(2x-7)^6}{12} + C$.  2. $u = x^2 + 4$: $(x^2+4)^3 + C$.  3. $-\dfrac{1}{2}e^{-x^2} + C$.  4. $u$ runs from 3 to 4: $\dfrac{1}{2}\left[\dfrac{u^4}{4}\right]_3^4 = \dfrac{256 - 81}{8} = \dfrac{175}{8} = 21.875$.

**By parts:** 1. $\dfrac{e^{3x}(3x - 1)}{9} + C$.  2. $\dfrac{x^2\ln x}{2} - \dfrac{x^2}{4} + C$.  3. $\big[-(x+1)e^{-x}\big]_0^1 = 1 - \dfrac{2}{e} \approx 0.264$.

**Partial fractions:** 1. $2\ln|x+2| + 5\ln|x+4| + C$.  2. $\dfrac{1}{(x-1)(x+4)} = \dfrac{1}{x-1} - \dfrac{1}{x+4}$ scaled: the integral is $\ln\left|\dfrac{x-1}{x+4}\right| + C$.  3. $\dfrac{1}{(x+1)(x+2)} = \dfrac{1}{x+1} - \dfrac{1}{x+2}$: $\big[\ln\tfrac{x+1}{x+2}\big]_0^1 = \ln\tfrac{2}{3} - \ln\tfrac{1}{2} = \ln\tfrac{4}{3}$.

**Separable DEs:** 1. $y = 3e^x$.  2. $N = 200e^{0.4t}$; $N(5) = 200e^2 \approx 1478$.  3. $T = 20 + 70e^{-0.1t}$; $40 = 20 + 70e^{-0.1t} \Rightarrow e^{-0.1t} = \tfrac{2}{7} \Rightarrow t = 10\ln\tfrac{7}{2} \approx 12.5$ minutes.  4. $B(t) = 40000 - 10000e^{0.05t}$; the pot empties when $e^{0.05t} = 4$, i.e. $t = 20\ln 4 \approx 27.7$ years. (Compare the steady state £40{,}000 of Section \@ref(constructing-des): starting below it, the pot is doomed — the solution says exactly when.)

**Interpretation:** 1. Total electricity used over the first 90 days, in kWh.  2. $V \to 0$ as $t \to \infty$; real machines retain scrap/salvage value, so the model undervalues old machines.  3. It puts the *fastest* growth at $t = 0$, whereas genuinely new products usually start slowly (awareness must build); the S-shaped **logistic** model captures that slow start.
